---
title: "Scale alignment across targets, engines and conventions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scale alignment across targets, engines and conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(scorecraft)
library(data.table)
```

## 1. Why "700 points" means nothing on its own

A points score is a linear function of a log-odds. Which log-odds, in
which orientation, and through which map, is what gives the number its
meaning. Two scorecards can both print "700" for a customer and disagree
completely about that customer's risk unless both raw scores were taken to
the **same scale** by the **same procedure**. That is the whole subject of
this vignette.

### The scale is one statement

`base_score`, `base_odds` and `pdo` are not three independent knobs; they
are a single sentence: *at `base_score` points the odds are `base_odds`,
and every `pdo` points they double*. From that sentence follow two
constants,

$$
\mathrm{factor} = \frac{\mathrm{pdo}}{\ln 2}, \qquad
\mathrm{offset} = \mathrm{base\_score} - \mathrm{factor} \cdot \ln(\mathrm{base\_odds}),
$$

and the map from log-odds to points,
$\mathrm{score} = \mathrm{offset} + \mathrm{factor} \cdot \ln(\mathrm{odds})$.
The textbook example is 600 points at odds 50:1 with a PDO of 20
(Siddiqi, 2006); it is an example, not a standard published by anyone.

```{r constants}
factor <- 20 / log(2)
offset <- 600 - factor * log(50)
c(factor = factor, offset = offset)
```

### The orientation trap

The word *odds* changes meaning between the two literatures. Under
`objective = "risk"` the event is the bad case, more points are safer, and
`base_odds` is **non-event:event** (`safe:event`). Under
`objective = "propensity"` the event is the good case, more points mean a
higher chance of the event, and `base_odds` is **event:non-event**
(`event:safe`). Feeding the same number `50` to both conventions is the
most common sign error in scorecard work, and it is silent: nothing fails,
the points simply mean something else.

`scr_align()` records `odds_orientation` on the object precisely so that
this can never be implicit. Let us see the trap on a simulated,
well-calibrated event logit.

```{r simulate}
set.seed(2026)
n   <- 4000
raw <- stats::rnorm(n, mean = stats::qlogis(0.12), sd = 1.2)   # event logit
y   <- stats::rbinom(n, 1, stats::plogis(raw))                 # calibrated outcome
```

With `method = "direct"` the model is trusted as calibrated: the log-odds
*is* the raw logit, in the orientation the direction implies (`S = -1`
under `higher_is_safer`, `+1` under `higher_is_riskier`, `I = 0`).

```{r direct}
al_risk <- scr_align(raw, y, base_score = 600, base_odds = 50, pdo = 20,
                     direction = "higher_is_safer", method = "direct")
al_risk
```

The same `50`, passed under the propensity convention, states that at 600
points the event is fifty times *more* likely than not. The equivalent
statement in that orientation is `base_odds = 1/50`.

```{r trap}
al_prop_wrong <- scr_align(raw, y, base_score = 600, base_odds = 50, pdo = 20,
                           direction = "higher_is_riskier", method = "direct")
al_prop_right <- scr_align(raw, y, base_score = 600, base_odds = 1 / 50, pdo = 20,
                           direction = "higher_is_riskier", method = "direct")

# what does "600 points" mean under each object? Invert score = a + b * raw
prob_at <- function(al, score) predict(al, (score - al$a) / al$b, type = "prob")
data.table(
  object      = c("risk 50 (safe:event)", "propensity 50 (event:safe)", "propensity 1/50 (event:safe)"),
  orientation = c(al_risk$odds_orientation, al_prop_wrong$odds_orientation, al_prop_right$odds_orientation),
  p_event_at_600 = round(c(prob_at(al_risk, 600), prob_at(al_prop_wrong, 600), prob_at(al_prop_right, 600)), 4)
)
```

Under the correct pair of statements the two scales are mirror images of
each other around 600, and both return the same event probability for the
same customer; under the wrong pair, "600 points" is a 2% customer on one
card and a 98% customer on the other.

```{r mirror}
head(data.table(
  raw   = round(raw, 3),
  p_true = round(stats::plogis(raw), 4),
  score_risk = round(predict(al_risk, raw), 1),
  score_prop = round(predict(al_prop_right, raw), 1),
  p_risk = round(predict(al_risk, raw, type = "prob"), 4),
  p_prop = round(predict(al_prop_right, raw, type = "prob"), 4)
), 5)
all.equal(predict(al_risk, raw) + predict(al_prop_right, raw), rep(1200, n))
```

Now the default, `method = "regression"`. On a calibrated logit it should
recover an intercept near 0 and a slope near `-1` (the sign of the
direction), and it does. The slope sits a little below 1 in magnitude
because the regression runs on band means: averaging the logit inside a
band is not the same as taking the logit of the band's event rate, and the
difference attenuates the slope slightly. That is a property of any banded
calibration, not a defect of this one.

```{r regression}
al_reg <- scr_align(raw, y, base_score = 600, base_odds = 50, pdo = 20,
                    direction = "higher_is_safer")
al_reg
```

## 2. The two-step regression alignment

`method = "regression"` does not trust the raw score to be a log-odds. It
measures how the raw score maps to the *empirical* log-odds and composes
that map with the PDO map:

1. band the raw score by quantiles on the reference data (`n_bands`,
   default 10);
2. in every band, compute the empirical log-odds with Laplace smoothing,
   in the orientation the direction implies;
3. regress the band log-odds on the band mean of the raw score, weighted by
   band size: $\ln(\mathrm{odds}) = I + S \cdot \mathrm{raw}$;
4. compose with the scale:
   $\mathrm{score} = \mathrm{offset} + \mathrm{factor}\,(I + S \cdot \mathrm{raw}) = a + b \cdot \mathrm{raw}$.

The bands are kept on the object, so the regression can be audited row by
row.

```{r bands}
al_reg$calibration$bands
c(intercept = al_reg$calibration$intercept, slope = al_reg$calibration$slope,
  adj_r2 = al_reg$calibration$r2)
```

How to read the three numbers:

* **intercept** $I$: the log-odds at `raw = 0`, in the scale's orientation.
  A value away from 0 means the raw score is shifted relative to the
  population it is being aligned on (a prior shift, a reweighted sample, a
  model fitted on a different base rate).
* **slope** $S$: how many log-odds units the empirical odds move per unit
  of raw score. Its **sign** must agree with the direction (`-1` expected
  under `higher_is_safer`, `+1` under `higher_is_riskier`); `scr_align()`
  warns when it does not. Its **magnitude** measures over- or
  under-confidence: $|S| < 1$ means the raw logit spreads more than the
  data support, $|S| > 1$ that it is too timid.
* **adj. R2**: how well a straight line explains the band log-odds. A low
  value means the raw score is not linear in log-odds, and a linear
  alignment will be systematically wrong in the tails; that is a signal to
  inspect the bands, not to trust the map.

Why not always use `"direct"`? Because raw scores are rarely calibrated on
the population they are deployed to. Shrink and shift the simulated logit
and compare the two methods:

```{r miscalibrated}
raw_bad <- 0.6 * raw - 1            # over-timid and shifted
al_bad_direct <- scr_align(raw_bad, y, method = "direct")
al_bad_reg    <- scr_align(raw_bad, y)
al_bad_reg$calibration[c("intercept", "slope", "r2")]
data.table(
  method = c("direct", "regression"),
  mean_predicted = c(mean(predict(al_bad_direct, raw_bad, type = "prob")),
                     mean(predict(al_bad_reg,    raw_bad, type = "prob"))),
  observed = mean(y)
)
```

The true map is $\ln(\mathrm{odds}_{\mathrm{safe:event}}) = -(1 + \mathrm{raw\_bad}) / 0.6$,
so intercept and slope should both be near $-1.67$. The regression lands at
about $-1.4$ and $-1.5$, attenuated by the banding as above but in the
right place, and its mean predicted probability matches the observed event
rate; the direct map reports whatever the raw score claims, here a third
too low. This is why `scr_scorecard()` always aligns and defaults to
`"regression"`.

## 3. Two targets, two conventions, one scale

The bundled `scr_demo` table carries two binary targets: `default`, a risk
target, and `churn`, which we shall treat as a propensity target (the
campaign wants to reach customers who *will* churn). Each is selected with
the other target dropped, so that neither becomes a candidate for the
other.

```{r select-two, results = "hide"}
cfg_risk <- scr_config(objective = "risk", verbose = FALSE, nthread = 1, use_ranger = FALSE,
                       use_lightgbm = FALSE, xgb_rounds = 60, n_boot = 20)
cfg_prop <- scr_config(objective = "propensity", verbose = FALSE, nthread = 1, use_ranger = FALSE,
                       use_lightgbm = FALSE, xgb_rounds = 60, n_boot = 20)

res_risk <- scr_select(scr_demo, "default", config = cfg_risk, drop = c("id", "churn"),
                       date_col = "ref_date")
res_prop <- scr_select(scr_demo, "churn", config = cfg_prop, drop = c("id", "default"),
                       date_col = "ref_date")

sc_risk <- scr_scorecard(res_risk)
sc_prop <- scr_scorecard(res_prop)
```

Both scorecards declare the same scale, 600/50/20, but the two alignment
objects differ in direction and in odds orientation, and therefore in the
sign of the map from logit to points.

```{r two-alignments}
sc_risk$alignment
sc_prop$alignment
c(risk = sc_risk$odds_orientation, propensity = sc_prop$odds_orientation)
```

`scr_score_metrics()` reports the AUC **in the direction of the scale**:
above 0.5 whenever the score ranks correctly in its own convention. The
`direction` column is carried on the table so the reader knows which way
the score was read.

```{r two-metrics}
scr_score_metrics(sc_risk)[, .(sample, direction, auc, auc_lo, auc_hi, ks)]
scr_score_metrics(sc_prop)[, .(sample, direction, auc, auc_lo, auc_hi, ks)]
```

Compare the mean score by outcome on hold-out. Under risk, events score
lower; under propensity, events score higher. The same points scale, read
the other way round.

```{r mean-by-outcome}
rbind(
  sc_risk$samples$holdout[, .(scorecard = "default (risk)", n = .N, mean_score = round(mean(score), 1)), by = y],
  sc_prop$samples$holdout[, .(scorecard = "churn (propensity)", n = .N, mean_score = round(mean(score), 1)), by = y]
)[order(scorecard, y)]
```

Reading a risk score as if it were a propensity score is the same error in
reverse: `scr_metrics()` with the wrong `higher_is_event` returns one
minus the AUC.

```{r wrong-way}
s <- sc_risk$samples$holdout
c(read_as_safer  = scr_metrics(s$score, s$y, higher_is_event = FALSE, ci = FALSE)$auc,
  read_as_riskier = scr_metrics(s$score, s$y, higher_is_event = TRUE,  ci = FALSE)$auc)
```

## 4. A tree challenger on the same scale

A gradient-boosted model can be fitted on the same WOE columns as the
scorecard and aligned to the same scale by the same regression. That makes
its output comparable point for point with the champion; it does **not**
make it a scorecard. The object says so explicitly: no points, no reason
codes, `supports_scorecard = FALSE`.

```{r challenger, results = "hide"}
sc_x <- scr_scorecard(res_risk, challenger = "xgboost")
```

```{r challenger-show}
sc_x$challenger$supports_scorecard
sc_x$challenger$points
sc_x$challenger$alignment
sc_x$challenger$metrics[, .(sample, direction, auc, auc_lo, auc_hi, ks)]
scr_score_metrics(sc_x)[sample == "holdout", .(sample, direction, auc, auc_lo, auc_hi, ks)]
```

The challenger's alignment has its own intercept and slope: two engines,
two raw scores, one declared scale. Here the slope is well above 1 in
magnitude, which says the boosted logit is compressed (a few dozen rounds
at a low learning rate leave it under-confident) and the alignment
stretches it back to the empirical odds; the scorecard logit, by contrast,
needed a slope close to 1. Its AUC interval overlaps the scorecard's,
which is the usual finding on a WOE feature set.

### The swap set

Aggregate AUCs hide where two models disagree. The swap-set table holds
the **approval rate** fixed and asks which customers one model would
approve and the other would not:

```{r swapset}
sc_x$challenger$swapset
```

How to read one row, say `approval_rate = 0.5`:

* both models approve the safest half of hold-out by their own score;
* `n_swap_in` customers are approved by the challenger but not by the
  champion; `n_swap_out` the reverse. At a fixed rate the two counts are
  equal;
* `event_rate_swap_in` versus `event_rate_swap_out` is the decision: the
  challenger is worth its complexity only if the customers it brings in are
  safer than the ones it sends out, by a margin that survives the sample
  size of the swap;
* `event_rate_champion` and `event_rate_challenger` are the event rates of
  the two approved books.

Because both scores sit on the same scale, the swap set can also be read
in points: a customer swapped in at the 70% rate crossed the challenger's
cut-off and not the champion's, and the two cut-offs are directly
comparable numbers.

## 5. The portfolio view

With several targets in flight, `scr_compare()` gives one row per target
(funnel, best hold-out model with its interval, warning signs) and
`scr_core()` lists the variables that survived on more than one target.

```{r portfolio}
runs <- list(default = res_risk, churn = res_prop)
scr_compare(runs)[, .(target, rows, event_rate, candidates, approved, best_model, auc, auc_lo, auc_hi, ks, relaxation)]
scr_core(runs, min_targets = 2)
```

A variable that is approved on both targets, with a good mean consensus
rank, is the strongest argument a model committee can hear in its favour:
it is not an artefact of one target definition. The `relaxation` column
records whether the consensus had to loosen its vote threshold to reach
the configured minimum number of variables, which is worth knowing before
the two shortlists are compared. Note that `scr_compare()`
reports the AUC of the **selection-stage** consensus models, not of the
final scorecards; the scorecards' own metrics live in
`scr_score_metrics()`, and since both cards are on the same scale, their
score distributions can be laid side by side as well.

```{r portfolio-scale}
rbind(
  data.table(scorecard = "default", sc_risk$samples$holdout[, .(min = min(score), median = stats::median(score), max = max(score))]),
  data.table(scorecard = "churn",   sc_prop$samples$holdout[, .(min = min(score), median = stats::median(score), max = max(score))])
)
```

## 6. Rescaling an existing scorecard

A committee may ask for a different convention, say 700 points at 30:1
with a PDO of 25. Refitting is unnecessary: the logit and the calibration
regression are unchanged, only the PDO map moves.

```{r rescale, results = "hide"}
sc_700 <- scr_scorecard(res_risk, base_score = 700, base_odds = 30, pdo = 25)
```

```{r rescale-show}
sc_700$alignment
# the calibration is the same; only factor and offset change
c(intercept_600 = sc_risk$alignment$calibration$intercept, intercept_700 = sc_700$alignment$calibration$intercept,
  slope_600 = sc_risk$alignment$calibration$slope, slope_700 = sc_700$alignment$calibration$slope)
```

Ranking metrics are invariant to a monotone transformation, so AUC, KS and
Gini are identical to the last digit, while every point value moves.

```{r rescale-metrics}
rbind(scr_score_metrics(sc_risk)[, .(scale = "600/50/20", sample, auc, ks, gini)],
      scr_score_metrics(sc_700)[, .(scale = "700/30/25", sample, auc, ks, gini)])
merge(sc_risk$points[, .(variable, bin, points_600 = points)],
      sc_700$points[, .(variable, bin, points_700 = points)],
      by = c("variable", "bin"))[1:6]
```

The two scores are the same log-odds seen through two maps, so one can be
recovered from the other exactly:

```{r rescale-identity}
a600 <- sc_risk$alignment; a700 <- sc_700$alignment
ln_odds <- (sc_risk$samples$holdout$score - a600$offset) / a600$factor
all.equal(a700$offset + a700$factor * ln_odds, sc_700$samples$holdout$score)
```

That identity is what "comparable" means: two scorecards are on the same
scale when the same log-odds gives the same points, and a scorecard is
rescaled, not remodelled, when only `factor` and `offset` change.

## 7. A checklist for the model committee

Everything the committee needs to reproduce, compare and later monitor the
scale is on the model card. Record it verbatim.

```{r model-card}
mc <- sc_risk$model_card
keys <- c("target", "objective", "direction", "odds_orientation",
          "base_score", "base_odds", "pdo", "factor", "offset",
          "align_method", "align_intercept", "align_slope", "align_r2",
          "n_train", "event_rate_train", "split_method", "split_cutoff",
          "challenger", "challenger_supports_scorecard", "auc_holdout", "ks_holdout")
data.table(field = keys, value = vapply(mc[keys], function(v) format(v, digits = 6), character(1)))
```

What to write down, and why:

1. **The scale as one sentence**: `base_score`, `base_odds`, `pdo`, and the
   derived `factor` and `offset`. Anyone with these five numbers can turn a
   log-odds into points and back.
2. **The orientation**: `objective`, `direction` and `odds_orientation`.
   Without them `base_odds = 50` is ambiguous; with them it is not.
3. **The alignment method and its coefficients**: `align_method`,
   `align_intercept`, `align_slope`, `align_r2`, and the number of bands.
   A `"regression"` alignment on one population is not the same map as a
   `"regression"` alignment on another; the coefficients are what make the
   two comparable, or show that they are not.
4. **The population the alignment was fitted on**: `n_train`,
   `event_rate_train`, the split method and cut-off. An alignment absorbs
   the base rate of that population; if the deployment population differs,
   the intercept will drift first.
5. **The challenger's status**: engine, `supports_scorecard = FALSE`, and
   the swap-set table at the approval rates the business uses.
6. **The invariants to check on every rescale**: AUC/KS/Gini unchanged,
   calibration intercept and slope unchanged, only `factor` and `offset`
   moved.

Two scorecards that share items 1 to 3 are comparable; two that share only
item 1 are not, however similar their numbers look.

## Reference

Siddiqi, N. (2006). *Credit Risk Scorecards: Developing and Implementing
Intelligent Credit Scoring*. Wiley.
