---
title: "Coarse classing with an audit trail"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coarse classing with an audit trail}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why manual coarse classing still matters

Optimal binning does one job well: it finds the cut points and the
groupings that maximise the information value of a variable subject to the
admission rules (minimum bin size, monotonicity, hold-out stability). What
it cannot know is how the business reads the variable. A bureau score is
communicated to underwriters in policy bands; a region is priced as north
and south, not as five states; an age is quoted in decades. A scorecard
whose bins cut at 33.36 and 48.06 is correct, but nobody in the credit
committee can explain it, and a bin nobody can explain is a bin nobody
will defend when the model is challenged.

There is also a stability argument. The optimal cut points are estimated
on one training window, and a handful of them sit on thin slices of the
distribution. Coarser, rounder bands lose a little information value on
train and often lose nothing on hold-out, while being far less likely to
drift when the population shifts. The lab in `scorecraft` lets the analyst
make exactly that trade, and shows the price of it on train and hold-out
before anything is committed.

What the lab refuses to do is let a manual decision enter silently. Every
proposal is compared against the optimal bins, every acceptance carries a
reason, every override of a blocking rule is a row in an append-only
ledger, and the automatic artefacts are frozen alongside the manual ones.
The scorecard, the R scoring function and the production SQL consume the
manual bins through the very same code path as the optimal ones, so there
is no second implementation to keep in step.

## A fast selection run

The lab opens on an `scr_select()` result. The configuration below is the
quick one used in the package tests: single thread, two consensus voters
and a light bootstrap, enough for a few seconds of run time on
`scr_demo`.

```{r setup}
library(scorecraft)
cfg <- scr_config(verbose = FALSE, nthread = 1, use_ranger = FALSE,
                  use_lightgbm = FALSE, xgb_rounds = 60, n_boot = 20)
res <- scr_select(scr_demo, "default", config = cfg,
                  drop = c("id", "churn"), date_col = "ref_date")
scr_selected(res)
```

## Opening the lab

`scr_coarse_classing()` covers every variable that reached binning, not
only the consensus shortlist, so a variable failed by screening can be
rebinned and forced in later with a reason. `author` is free text recorded
in the ledger; it defaults to the system user.

```{r open}
lab <- scr_coarse_classing(res, author = "analyst")
lab
```

Without a variable, `scr_classing_view()` prints one line per variable:
its current source (optimal or manual), the number of bins, the train and
hold-out IV, the PSI and a verdict computed with the same rules a proposal
will face. The last column says whether the variable is currently in the
final shortlist (`yes`, or `-`).

```{r overview}
ov <- scr_classing_view(lab)
```

With a variable, the view is the bin table with train and hold-out side
by side, plus a text bar chart of the event rate. `vl_score_01` is a
numeric with seven optimal bins; `ds_regiao` is a categorical with one
bin per state.

```{r view-num}
scr_classing_view(lab, "vl_score_01")
```

```{r view-cat}
scr_classing_view(lab, "ds_regiao")
```

## Proposing bins

`scr_classing_propose()` takes exactly one instruction per call. For a
numeric variable the instruction is `breaks` (absolute interior cut
points), `merge` (adjacent bin ids) or `split` (`c(id, at)`). For a
categorical it is `groups` (a list of character vectors, optionally
named), `merge` (bin ids), `missing_to` (which bin receives the
`"MISSING"` category) or `other_to` (the catch-all bin for every training
category not listed). A proposal is a value: nothing changes in the lab
until it is accepted.

### Numeric: breaks, merge, split

Suppose underwriting quotes `vl_score_01` in the bands below 40, 40 to
55, 55 to 70 and above 70.

```{r propose-breaks}
p_breaks <- scr_classing_propose(lab, "vl_score_01", breaks = c(40, 55, 70))
p_breaks
```

The header repeats the instruction. The comparison table has one row per
metric and three columns: the optimal bins, the proposal and the
difference. Here the four policy bands cost about 0.05 of IV on train and
about 0.014 on hold-out, while the IV ratio (hold-out over train), the KS
and the smallest bin all improve, and the PSI halves. Below the table are
the manual bins themselves with count, share, event rate and WOE on train
and on hold-out. No warning was raised, so the verdict is `ACCEPTABLE`.

`merge` and `split` are relative to the *current* bins of the variable
and resolve to absolute cut points, which is what the header shows.

```{r propose-merge-split}
p_merge <- scr_classing_propose(lab, "vl_score_01", merge = c(1, 2))
p_merge$entry$cutpoints
p_split <- scr_classing_propose(lab, "vl_score_01",
                                split = c(1, res$fit$results$vl_score_01$cutpoints[1] - 5))
p_split
```

The split proposal is a `REVIEW`: the new first bin holds 2% of the
training rows and breaks the monotone pattern of the event rate, so the
engine's `NOT_MONOTONIC` rule fires. A `REVIEW` verdict is advisory: the
proposal can be accepted with a reason or discarded. Every proposal takes
the next id from the lab's counter (`P001`, `P002`, ...), whether or not
it is later acted on, so the ids in a session count the proposals made,
not the decisions taken. A proposal is a value and can be kept for as long
as the lab is open; the lab only refuses a proposal that has already been
accepted or discarded.

### Categorical: groups, other_to, missing_to

Pricing works with two regions. Names given to the groups are display
labels; the bin label stored in the entry is the categories joined by the
configuration's separator, exactly as the binning engine writes it.

```{r propose-groups}
p_groups <- scr_classing_propose(lab, "ds_regiao",
                                 groups = list(south = c("BA", "RS"),
                                               north = c("SP", "RJ", "MG")))
p_groups
```

Every training category must land somewhere. Listing them all is one
option; the other is to name a catch-all bin with `other_to`, which is
also how a reviewer says "everything else goes here". The result is the
same grouping, and the entry records that the second bin is the catch-all
(`is_other`).

```{r propose-other}
p_other <- scr_classing_propose(lab, "ds_regiao",
                                groups = list(south = c("BA", "RS"), rest = "SP"),
                                other_to = "rest")
p_other$entry$bin
p_other$entry$manual$is_other
```

`ds_optin` carries a `"MISSING"` category, which the optimal binning kept
on its own. `missing_to` folds it into another bin.

```{r propose-missing}
p_missing <- scr_classing_propose(lab, "ds_optin", missing_to = 1)
p_missing
```

### Reading the warnings and the verdict

Every proposal is screened with the eight engine rules on train,
revalidated on hold-out with the bins frozen, and checked against a few
lab-specific rules. Codes fall into two tiers.

* **Warnings** (advisory) give a `REVIEW` verdict: the engine screening
  codes (`NOT_MONOTONIC`, `SMALL_BIN`, `IV_BELOW_MIN`, `IV_SUSPECT`, ...),
  the hold-out codes (`IV_DROPS_ON_HOLDOUT`, `IV_LOW_ON_HOLDOUT`,
  `PSI_UNSTABLE`, ...) and `IV_LOSS_VS_OPTIMAL`, raised when the hold-out
  IV falls more than `lab_max_iv_loss` (10% by default) below the optimal
  one. The `ds_regiao` grouping is a `REVIEW` for exactly that reason: two
  regions lose about a fifth of the hold-out IV of five states.
* **Blocking** codes give a `BLOCKED` verdict: an empty bin, a degenerate
  bin (no events or no non-events, unless the lab was opened with
  `laplace > 0`), a bin below `lab_min_bin_pct_hard` (0.5%) and a manual IV
  crossing `iv_max`, the leakage ceiling. The lab must not be the place
  where leakage is manufactured.

`ACCEPTABLE` means no code at all. The verdict, the codes and the reason
travel with the decision into the ledger.

## Accepting, discarding and the blocked path

`reason` is mandatory, at least five characters, and it is the audit
trail: write what a reviewer would need to read a year from now. Both
verbs return the updated lab, so the idiom is to reassign.

```{r accept}
lab <- scr_classing_accept(lab, p_breaks, reason = "policy bands 40/55/70 used by underwriting")
```

The grouping proposed in the previous section is still valid and is
accepted in turn.

```{r accept-groups}
lab <- scr_classing_accept(lab, p_groups, reason = "north/south is what pricing uses")
```

The `ds_optin` proposal erased what little signal the variable had
(`IV_BELOW_MIN`), so it is discarded, with a reason, and the variable
keeps its optimal bins. A discarded proposal is a ledger row too.

```{r discard}
p_missing <- scr_classing_propose(lab, "ds_optin", missing_to = 1)
lab <- scr_classing_discard(lab, p_missing, reason = "folding MISSING into SIM erases the signal")
```

A blocking rule cannot be accepted through the normal path. A break at
-5000 on `vl_score_04`, whose minimum is above 16, leaves the first bin
empty:

```{r blocked}
p_blocked <- scr_classing_propose(lab, "vl_score_04", breaks = c(-5000, 50))
p_blocked$verdict
p_blocked$blocking
```

```{r blocked-accept, error = TRUE}
scr_classing_accept(lab, p_blocked, reason = "we need this band for the policy")
```

The override path is `override = TRUE`. It exists because there are
legitimate reasons to carry a band the training data does not populate (a
policy floor that will bind on a future population, for example), but it
is never silent: the ledger receives an `override` row naming the blocking
codes, followed by the `accept` row. Here it is exercised on a copy of the
lab so that the session carries on without the empty bin.

```{r override}
lab_override <- scr_classing_accept(lab, p_blocked, reason = "deliberate policy floor at -5000",
                                    override = TRUE)
scr_decisions(lab_override)[variable == "vl_score_04", .(seq, action, proposal_id, verdict, warnings, reason)]
```

## Choosing the variables

`scr_classing_choose()` builds the final list as
`(consensus shortlist + force) - drop`, optionally intersected with
`keep`. `force` is allowed for any variable that reached binning, with two
exceptions that need `override = TRUE`: a variable failed for
`IV_SUSPICIOUS` (the leakage ceiling) and a derived `__sp` flag when
`allow_derived_final = FALSE`. `reason` is one string for every variable
named, or a character vector named by variable.

`vl_score_10` is in the consensus shortlist but will not be available at
decision time; `vl_score_03` was failed by screening for `NOT_MONOTONIC`
but policy requires it on the card.

```{r choose}
scr_funnel(res, cols = "all")[feature %in% c("vl_score_10", "vl_score_03"),
                              .(feature, exit_stage, screen_reason)]
lab <- scr_classing_choose(lab, drop = "vl_score_10", force = "vl_score_03",
                           reason = c(vl_score_10 = "not available at decision time",
                                      vl_score_03 = "policy: bureau band must be scored"))
```

## The session summary and the ledger

Printing the lab summarises the session: the variables touched with bins
and IV before and after, the verdicts, the reasons, the discards and the
final choice.

```{r summary}
lab
```

`scr_decisions()` returns the ledger: one row per decision, append-only,
with the author, the timestamp, the instruction, the metrics before and
after, the verdict, the codes and the reason. The same function reads the
ledger from the committed result and from the scorecard fitted on it.

```{r ledger}
scr_decisions(lab)[, .(seq, variable, action, proposal_id, verdict, reason)]
```

## The spec round trip: a reviewer edits in a spreadsheet

Not every reviewer works in R. `scr_classing_spec()` writes the classing
as a long table, one row per bin of every variable, optimal and manual.
The authoritative columns a reviewer may edit are `lower` and `upper` for
numerics, `categories` and `is_other` for categoricals, and `reason`;
everything else (counts, rates, WOE) is context and is regenerated on
read. Open ends are written as empty cells. A `.xlsx` path writes a
workbook through openxlsx; a `.csv` path needs nothing.

```{r spec}
spec <- scr_classing_spec(lab)
spec
spec_file <- file.path(tempdir(), "classing_default.csv")
scr_classing_spec(lab, file = spec_file)
```

A reviewer opens the file, moves the first cut of `vl_score_01` from 40 to
42 and writes why in `reason`. Done in R, as it would be done in a
spreadsheet cell by cell:

```{r edit}
sheet <- read.csv(spec_file, stringsAsFactors = FALSE)
i1 <- sheet$variable == "vl_score_01" & sheet$bin_id == 1
sheet$upper[i1]  <- 42
sheet$reason[sheet$variable == "vl_score_01"] <- "reviewer: first cut moved to 42 to match the bureau band"
write.csv(sheet, spec_file, row.names = FALSE, na = "")
```

`scr_classing_read()` validates the file before anything else happens:
`bin_id` must run from 1 to k without gaps, `upper` must be finite and
strictly increasing except on the last bin, and the `lower` of every bin
must equal the `upper` of the one before it. The edit above touched only
`upper`, so the file is refused with a named reason.

```{r read-refused, error = TRUE}
scr_classing_read(spec_file)
```

Fixing the neighbouring `lower` makes the spec contiguous again.

```{r read-ok}
sheet$lower[sheet$variable == "vl_score_01" & sheet$bin_id == 2] <- 42
write.csv(sheet, spec_file, row.names = FALSE, na = "")
spec_back <- scr_classing_read(spec_file)
```

`scr_classing_import()` compares the file with the lab's current bins and
turns every variable that differs into a proposal, with the same checks
and the same comparison as a proposal made by hand. The reviewer's reason
arrives as `imported_reason`; nothing is accepted on the analyst's behalf.

```{r import}
imported <- scr_classing_import(lab, spec_back)
names(imported)
imported$vl_score_01$imported_reason
imported$vl_score_01
```

The comparison now has a fourth column, `current`, because the variable
already carries an accepted manual proposal. Accepting the imported one
supersedes it, and the ledger records both facts.

```{r import-accept}
lab <- scr_classing_accept(lab, imported$vl_score_01, reason = imported$vl_score_01$imported_reason)
scr_decisions(lab)[variable == "vl_score_01", .(seq, action, proposal_id, instruction, verdict)]
```

## Committing the lab: what changed

`scr_classing_apply()` returns a new `scr_result`. The input result is
not modified. Inside the new one, the accepted manual entries replace the
optimal ones in `fit`, the screening and hold-out rows of those variables
are recomputed with the same pipeline functions, the final shortlist is
the one implied by the choice, and the funnel, gains, SQL and summary are
rebuilt.

```{r apply}
res2 <- scr_classing_apply(lab)
res2
```

`scr_selected()` on the new result returns the final list; the automatic
one is still there under `which = "consensus"`.

```{r selected}
scr_selected(res2)
scr_selected(res2, "consensus")
setdiff(scr_selected(res2), scr_selected(res2, "consensus"))
setdiff(scr_selected(res2, "consensus"), scr_selected(res2))
```

The funnel gains two columns. `provenance` says what the lab did to each
variable (`auto`, `manual:rebin`, `manual:add`, `manual:drop`, or
`manual:rebin+add` / `manual:rebin+drop` when both happened) and
`manual_reason` carries the reason. A dropped variable exits at a new
stage, `08.manual_drop`; a forced one is `07.approved` even though the
consensus never selected it.

```{r funnel}
touched <- c("vl_score_01", "ds_regiao", "vl_score_03", "vl_score_10")
scr_funnel(res2, cols = "all")[feature %in% touched,
                               .(feature, exit_stage, provenance, manual_reason)]
```

The automatic fit is frozen as `fit_auto`, so the optimal cut points
remain available next to the manual ones for as long as the result lives.

```{r frozen}
res2$fit_auto$results$vl_score_01$cutpoints
res2$fit$results$vl_score_01$cutpoints
res2$fit$summary[res2$fit$summary$feature %in% c("vl_score_01", "ds_regiao"),
                 c("feature", "algorithm", "n_bins", "total_iv")]
```

## Refitting the scorecard

`scr_scorecard()` on the committed result fits the logistic regression on
the WOE columns of the final list, the manual ones included, and aligns
the score exactly as before. The points of a manually binned variable are
distributed over its four policy bands.

```{r scorecard}
sc <- scr_scorecard(res2)
sc
sc$points[variable == "vl_score_01", .(variable, bin, woe, points)]
sc$points[variable == "ds_regiao", .(variable, bin, woe, points)]
```

The model card states the provenance in words: which binning algorithms
the card mixes, whether the shortlist came from the consensus or from the
lab, how many manual bins it carries and which variables were forced in
or dropped. The ledger travels into the scorecard as well.

```{r model-card}
str(sc$model_card[c("binning_algorithm", "shortlist_source", "n_manual_bins",
                    "manual_bins", "forced_in", "manual_dropped", "n_decisions")])
nrow(scr_decisions(sc))
```

## Production: R and SQL follow the manual bins

Nothing downstream needs to know that a bin was drawn by hand.
`scr_apply()` scores new rows in R with the frozen pre-processing and the
frozen bins; the points for `vl_score_01` fall in one of the four policy
bands and the points for `ds_regiao` in one of the two regions.

```{r apply-new}
new <- head(scr_demo, 5)
new[, c("vl_score_01", "ds_regiao")]
scr_apply(sc, new, what = "points")[, .(score, score_points, vl_score_01_points, ds_regiao_points)]
```

`scr_sql()` emits the same thing for the database. The header carries the
provenance line; for the two variables touched, the lines below show the
frozen pre-processing, the WOE `CASE` on the manual cut points and
groupings with full precision, and the bin-index `CASE` on the same cuts.
The tail composes the whole points from that index.

```{r sql}
sql <- scr_sql(sc, table = "prd.customers", dialect = "databricks")
sql_lines <- unlist(strsplit(sql, "\n", fixed = TRUE))
cat(grep("^-- Provenance", sql_lines, value = TRUE), sep = "\n")
cat(grep("WHEN (vl_score_01|ds_regiao) ", sql_lines, value = TRUE), sep = "\n")
cat(tail(sql, 8), sep = "\n")
```

That the two paths agree number for number, including a value sitting
exactly on a manual cut point, is verified by the package tests, which run
the generated SQL in DuckDB and compare it with `scr_apply()` on the same
rows. This vignette does not need a database to run.

## Governance

The lab is designed so that a reviewer can reconstruct every manual
decision from the deliverables alone.

* **What the ledger records.** One row per decision, in order:
  `accept`, `discard`, `supersede`, `restore` (a `reset = TRUE` proposal
  taking a variable back to its optimal bins), `override`, `force`,
  `drop` and `keep`. Each row carries the author, the timestamp, the
  instruction as typed, the number of bins and the train and hold-out IV
  before and after, the PSI, the verdict, the codes and the reason. The
  ledger is append-only and travels unchanged from the lab into the
  result, into the scorecard and into the `.xlsx` workbooks written by
  `scr_export()`, next to the classing spec.
* **What is blocked.** An empty bin, a degenerate bin without smoothing, a
  bin below the hard minimum share, a manual IV above the leakage ceiling,
  a category left unassigned, a proposal without a reason, a `force` of a
  variable that never reached binning. The first four block a proposal
  (`BLOCKED`); the last three are errors that never produce a proposal at
  all.
* **What needs an override.** Accepting a `BLOCKED` proposal, forcing a
  variable failed for `IV_SUSPICIOUS`, and forcing a derived `__sp` flag
  under `allow_derived_final = FALSE`. An override is not a way round the
  rule; it is a documented exception, with its own ledger row, that the
  model card and the SQL header will point to.

Everything a manual decision touches remains reproducible: the optimal
artefacts are frozen alongside, the manual bins are recomputed on the
training rows with the engine's own WOE formula, and the scorecard, the
R scoring and the SQL read them through the same contract as any optimal
bin.
