---
title: "LGD and EAD under IRB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGD and EAD under IRB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_duckdb <- requireNamespace("duckdb", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)
has_openxlsx <- requireNamespace("openxlsx", quietly = TRUE)
```

The two IRB parameters after the PD follow the discipline of the scorecard
pipeline: a reference data set built with a funnel that names every rule
it applied, drivers binned on training cohorts and revalidated with frozen
bins on a hold-out, pools, a downturn and a floor that land in a ledger
with a reason, an R scoring function and a SQL query that agree number for
number, a validation battery with traffic lights and a workbook. Part A
walks the loss given default (LGD) on the bundled default events and cash
flows; part B walks the exposure at default (EAD) on the bundled facility
snapshots. Both use the light configuration below: single thread, twenty
bootstrap replicates, and a lower minimum of defaults per CCF bin because
the EAD panel is small.

```{r setup}
library(scorecraft)
cfg <- scr_config(verbose = FALSE, nthread = 1, n_boot = 20, ccf_min_defaults = 20)
scr_verbose(FALSE)
```

# Part A: loss given default

## What a workout LGD is

A workout LGD is the economic loss of a default event over the exposure at
the default date: the exposure, less the recoveries, plus the direct
costs, the drawings after default and the indirect costs allocated to the
event, every cash flow discounted to the default date at the reference
rate in force at that date plus an add-on. A cure (the facility returns to
performing) is not a zero loss by decree: the balance outstanding at the
cure date enters as a recovery on that date, so the cure carries its costs
and the discount effect. Two defaults of one facility closer than a cure
window are one event. An incomplete workout (still open at the observation
date) receives the expected further recovery read from the recovery
profile of the closed defaults of the same product; an open event older
than the maximum workout period is closed with no further recovery. The
long-run average (LRA) is the default-weighted mean over the events; the
downturn LGD is the value appropriate for an economic downturn, never
below the LRA; and an own estimate is subject to an input floor by asset
class and collateral. The `"type1"` and `"type3"` labels of the downturn
methods below follow the public supervisory guidance on downturn LGD
estimation (observed impact, and reference value plus add-on); the `"bcb"`
parameter preset is the Brazilian text (Resolução BCB 303/2023).

## The reference data set: `scr_workout()`

`scr_demo_lgd` holds 900 default events on three products, with the
drivers a workout model reads; `scr_demo_lgd_cashflows` is the long table
of their post-default cash flows; `scr_demo_rates` the monthly reference
rate. `scr_workout()` discounts, merges, extrapolates and bounds, and
returns the reference data set (RDS) with one row per event.

```{r workout}
wo <- scr_workout(scr_demo_lgd, scr_demo_lgd_cashflows, rates = scr_demo_rates, config = cfg)
wo
```

The funnel is the first deliverable: the thirty second defaults of the
demo were fifteen pairs closer than the nine-month cure window, merged
into their first spell, plus fifteen pairs far enough apart to stay two
events; two open events older than sixty months were closed with no
further recovery; 160 open events younger than that received an expected
further recovery; three events lost more than the exposure (costs on a
facility with no recovery) and were kept above one because
`lgd_cap_at_one` is off.

```{r funnel}
wo$funnel[, .(rule, n, action)]
```

The long-run average is reported default-weighted, the regulatory
estimate, and exposure-weighted. They differ by ten points here because
the mortgages carry the exposure and the low LGD. The share of incomplete
workouts is the share of rows whose loss is partly an extrapolation.

```{r lra}
wo$summary$by_product
wo$summary$share_incomplete
```

The recovery profile is the cumulative discounted recovery rate of the
closed defaults by product and month in default; it is what the
extrapolation reads (`rho_tau` at the current age, `rho_t_max` at the
end), and what the in-default grid will read later.

```{r profile}
wo$recovery_profile[month %in% c(6, 12, 24, 36, 60)]
head(wo$extrapolation[, .(default_id, product, months_in_default, rho_tau, rho_t_max, expected_further)], 4)
```

## Binning a bounded target: `scr_bin_continuous()`

The optimal-binning engine of the scorecard accepts binary targets only.
LGD and CCF are continuous and bounded, so the severity stage and the CCF
pools use a supervised binner of their own: candidate cut points at
quantiles, greedy merges of the adjacent pair whose merge loses the least
between-bin sum of squares until the target number of bins and the size
floors hold, then pool-adjacent violators on the bin means when
monotonicity is required. The result has exactly the shape of the engine's
object: the `woe` slot carries the bin mean of the target and the `iv`
slot the bin's share of the between-bin sum of squares, so the total is
the eta-squared of the driver and the same SQL emitter reproduces the bin
statistic in every dialect. Here it runs once, on the worst delinquency
before default, over the non-cured events, with the 2024 and later
defaults as the hold-out.

```{r cbin}
nc <- wo$rds[is_cure == FALSE]
cb <- scr_bin_continuous(nc, "lgd_real", "prior_dpd_max",
                         train_idx = which(nc$year < 2024), holdout_idx = which(nc$year >= 2024))
cb
cb$fit$results$prior_dpd_max$bin
cb$fit$results$prior_dpd_max$woe
```

The hold-out revalidation recomputes the bin means with the bins frozen,
reports the PSI of the bin shares with the sample-size-adjusted critical
value, and flags `UNSTABLE_HOLDOUT` when the hold-out means break the
training order.

```{r cbin-holdout}
cb$holdout[, .(bin, n_train, mean_train, n_holdout, mean_holdout)]
cb$summary[, .(feature, eta2, eta2_holdout, psi, psi_flag, holdout_ok)]
```

## The two-stage model: `scr_lgd()`

`scr_lgd()` fits `LGD = P(cure) LGD_cure + (1 - P(cure)) E[LGD | no cure]`.
The cure stage is a binary model on `is_cure` with the scorecard
machinery: optimal binning on the training cohorts, WOE, hold-out
revalidation, a logistic regression on the WOE columns with the sign
check. The severity stage bins the same drivers against the realised LGD
of the non-cures with the continuous binner and fits a fractional logit
on the bin means. The split is by cohort of default: the last 30% of the
default dates are the hold-out.

```{r lgd}
drv <- c("product", "ltv", "prior_dpd_max", "months_on_book", "region")
m <- scr_lgd(wo, drivers = drv, config = cfg)
m
```

Each stage keeps its own audit trail. In the cure stage, a driver enters
the regression only if its training IV clears the minimum, its frozen
bins hold on the hold-out and its coefficient is positive.

```{r cure}
m$cure$bins[, .(iv = sum(iv), iv_holdout = iv_holdout[1], holdout_ok = holdout_ok[1],
                selected = selected[1]), by = variable]
m$cure$sign_check
```

In the severity stage, `ltv` is removed by the sign check. Its bin means
fall with the loan-to-value because the ratio is zero for unsecured
facilities, which are the ones that lose most: the driver is the product
effect in disguise, and once `product` is in the regression its
coefficient turns negative.

```{r severity}
m$severity$bins[, .(variable, bin, count, mean, mean_holdout, selected)]
m$severity$sign_check
```

Metrics on both samples come with a bootstrap interval on Somers' D (the
generalised AUC is `(D + 1) / 2`) and on the loss capture ratio.

```{r metrics}
m$metrics[, .(sample, n, rmse, r2, spearman, gauc, gauc_lo, gauc_hi, lcr)]
```

The pools cut the training predictions into quantile bands, merge the
bands with fewer than `lgd_min_defaults_bin` defaults into the neighbour
with the closer long-run average, then merge adjacent bands whose
averages break the increasing order. With 620 training defaults and a
minimum of 100 per pool, the seven target bands collapse to three. Per
pool: the default-weighted LRA, the exposure-weighted one, and the
category-C margin of conservatism, a one-sided 95% t interval on the
mean. The downturn column is provisional at this point (a type-3 add-on
by configuration) and the floor is zero until the next two calls run.

```{r pools}
m$pools[, .(pool, pred_lo, pred_hi, n, lra, lra_ew, se, moc_c, lgd_dt, floor, lgd_final)]
```

## Downturn, floor and the in-default grid

`scr_lgd_downturn()` quantifies the downturn per pool from user-supplied
periods. Under `"type1"` the downturn value is the realised LGD of the
defaults whose default date falls in the periods (a pool with fewer than
ten such defaults falls back to the add-on); the reference value, the mean
of the two worst calendar years of the pool, is reported as a challenger,
not a bound. The downturn LGD used for capital is
`min(1, max(LRA + MoC, DT + MoC))`. The reference rate of the demo rises
above thirteen per cent in 2022 and 2023, which is the reason recorded.

```{r downturn}
m <- scr_lgd_downturn(m, periods = data.frame(start = as.Date("2022-01-01"), end = as.Date("2023-12-31")),
                      reason = "reference rate above 13% in 2022-2023")
m$downturn$table[, .(pool, n, lra, moc_c, n_downturn, dt_observed, reference_value, method_used, lgd_dt, impact, below_reference)]
```

Pools 2 and 3 sit below their reference value: the two worst years of
those pools were worse than the periods declared. The flag does not change
the estimate; it is there for the reviewer to answer.

`scr_lgd_floor()` reads the input floor from the parameter table of the
framework. The tables are numbers, editable and printable; an edit is
detected and recorded as `params_modified` in the ledger. The floor of a
partly secured pool blends the unsecured and the collateral floors with
the secured share: `0.30 * 0.6 + 0.10 * 0.4 = 0.22` here.

```{r floor}
p <- scr_irb_params("bcb")
p$lgd_floor
m <- scr_lgd_floor(m, params = p, asset_class = "retail_other", secured_share = 0.4)
m$floors$table
m
```

The floor binds in none of the three pools on this data, and that fact is
what the ledger records. `scr_elbe()` derives the in-default grid: for
every pool and every age since default, the expected loss best estimate
(ELBE) is the mean realised LGD of the training defaults still in workout
at that age, and the in-default LGD adds the unexpected-loss increment
`(LGD_DT - LRA)` scaled by the share of the recoveries still to come,
read from the recovery profile. At age zero the ELBE equals the LRA and
the in-default LGD equals the downturn LGD; the consistency table checks
exactly that. The jump at six months is the cures leaving the set of open
events.

```{r elbe}
e <- scr_elbe(m)
e
e$consistency
```

## Production: `scr_apply()` and `scr_sql()`

`scr_apply()` scores new rows with the frozen bins of both stages and
returns the cure probability, the severity, the predicted LGD, the pool
and the pool parameters.

```{r apply}
new <- head(scr_demo_lgd, 5)
ap <- scr_apply(m, new, what = "all")
ap
```

`scr_sql()` emits the same computation for the database: the bin
statistics of both stages, the two logits, the predicted LGD, the pool
`CASE` on the pool boundaries (snapped to the midpoint between adjacent
predictions, so that floating-point noise cannot flip a pool between R
and SQL), the pool parameters and the floored result.

```{r sql}
sql <- scr_sql(m, table = "prd.defaults", dialect = "duckdb")
sql_lines <- unlist(strsplit(sql, "\n", fixed = TRUE))
cat(grep("AS pool$|AS lgd_dt,$|AS lgd_floor$|AS lgd_final$", sql_lines, value = TRUE), sep = "\n")
```

When `duckdb` is installed the query runs on the same five rows and
reproduces `scr_apply()`.

```{r sql-duckdb, eval = has_duckdb}
con <- suppressMessages(DBI::dbConnect(duckdb::duckdb(), config = list(threads = "1")))
DBI::dbWriteTable(con, "defaults_t", new)
got <- DBI::dbGetQuery(con, paste(scr_sql(m, table = "defaults_t", dialect = "duckdb"), collapse = "\n"))
DBI::dbDisconnect(con, shutdown = TRUE)
got[, c("lgd_pred", "pool", "lgd_dt", "lgd_floor", "lgd_final")]
identical(as.integer(got$pool), ap$pool)
all.equal(got$lgd_final, ap$lgd_final)
```

## Validation and export

`scr_lgd_validate()` runs on the hold-out (or on new data) against the
training reference: calibration per pool and for the portfolio (one-sided
t-test where under-estimation is the failure, loss shortfall, coverage of
the realised mean by the downturn LGD), discrimination against the
training gAUC, stability of the pool and driver-bin distributions, and
the homogeneity within and heterogeneity between pools. The lights use the
p-value thresholds of `pd_lights` and the fixed PSI thresholds.

```{r validate}
v <- scr_lgd_validate(m)
v
v$calibration[, .(pool, n, lgd_est, real_mean, real_ew, p, light, dt_covers)]
```

The red light on homogeneity is the price of three pools: a Welch test
between the two halves of each pool, split at its median prediction, finds
that the halves still differ, so the pools are not yet homogeneous. That is
what the battery is for; the fix is more defaults or a lower minimum per
pool, and either is a ledger row.

`scr_export()` writes one workbook with the RDS funnel and summaries, the
recovery profile, the bins and coefficients of both stages, the sign
checks, the pools, the downturn and its periods, the floors, the
in-default grid, the validation blocks, the model card and the ledger,
plus the SQL file.

```{r export, eval = has_openxlsx}
out <- file.path(tempdir(), "irb-vignette")
basename(unlist(scr_export(m, out, stamp = FALSE, validation = v, elbe = e)$files))
```

# Part B: exposure at default

## Realised conversion factors

Write `R` for the reference date, `D` for the default date, `E` for the
drawn amount and `L` for the limit. The realised credit conversion factor
of a default event is the share of the undrawn amount at the reference
date that was drawn by the default date,
`CCF = (E_D - E_R) / (L_R - E_R)`, and the realised EAD is the drawn
amount at the default date, never capped at the limit. The reference date
is one horizon before the default (twelve months here under the fixed
approach; the start of the calendar cohort under the cohort approach). The
CCF is only meaningful while the undrawn part is material: when the
utilisation at the reference date is at or above `ccf_u_star`, when
nothing is undrawn, or when the facility is over its limit, the
denominator is small or negative and the row is routed to the limit factor
`LF = E_D / L_R` instead of being dropped; a facility with no usable
limit uses the exposure factor `E_D / E_R`. Two floors apply at different
places: the realised CCF is floored at zero for the averages (a facility
that repaid before default; the raw value is kept), and the applied CCF of
an own estimate is floored at a fraction of the standardised CCF, one half
of `ccf_sa_ccf` here, while the predicted EAD is never below the drawn
amount.

## The reference data set: `scr_ead_data()`

`scr_demo_ead` is a panel of 1,200 revolving facilities over thirty
months with a 0/1 default flag; every run of ones opens a default event at
its first month. `scr_ead_data()` finds the reference snapshot of every
event, computes the realised measure and applies the funnel rules.

```{r ead-data}
ed <- scr_ead_data(scr_demo_ead, facility_id = "facility_id", obligor_id = "obligor_id",
                   date_col = "ref_date", limit = "limit", drawn = "drawn", defaulted = "defaulted",
                   drivers = c("product", "months_on_book", "dpd"), config = cfg)
ed
```

The funnel keeps everything it can and names what it did: a facility
younger than the horizon is kept from its first snapshot and flagged
`FAST_DEFAULT`; an over-limit facility is routed to the limit factor;
a negative realised CCF (the facility repaid before defaulting) is floored
at zero; a CCF above one (a limit raised or a drawing beyond the limit) is
kept as observed because no cap is configured.

```{r ead-funnel}
ed$funnel
head(ed$rds[, .(facility_id, ref_date, default_date, utilisation_ref, ead_realised, measure, ccf_raw, ccf, rule)])
```

The summary gives the simple and the exposure-weighted averages by
cohort of reference date and by measure, with a total row. The large
lines draw a smaller share of their undrawn amount, so the
exposure-weighted CCF sits below the simple one. Two cohorts and 1.4
years of reference dates is not a long run yet, and the print says so.

```{r ead-summary}
ed$summary
```

## Pools and downturn: `scr_ead()` and `scr_ead_downturn()`

`scr_ead()` splits by reference date (the most recent dates are the
hold-out), bins every candidate driver against the realised CCF on the
training rows with the continuous binner, revalidates the frozen bins on
the hold-out, and admits a driver only when it passes four named rules:
enough defaults in every bin, separation of the bin means (an F-test),
the training order preserved on the hold-out, and a stable bin
distribution. The cells of the cross of the admitted drivers are ordered
by predicted CCF and merged into pools with at least `ccf_min_defaults`
each; rows in the limit-factor measure form their own pool `LF`.

```{r ead}
m_ead <- scr_ead(ed, drivers = c("utilisation_ref", "product", "months_on_book", "dpd"), config = cfg)
m_ead
m_ead$drivers[, .(feature, n_bins, eta2, direction, p_anova, eta2_holdout, psi_flag, admitted, reason)]
```

Only the utilisation at the reference date is admitted: the product does
not separate the means, the months on book reverse their order on the
hold-out and the days past due collapse to one bin. Two pools follow the
two utilisation bins, the lower utilisation drawing the larger share of
its undrawn amount. Per pool: the long-run average, `moc_est`, a
one-sided normal estimation-error margin at `ccf_moc_alpha`;
`ccf_final = max(lra, ccf_dt) + moc_est`; the floor
`0.5 * 0.40 = 0.20`; and `ccf_applied = max(ccf_final, ccf_floor)`. The
`LF` pool has no scalar floor because its floor depends on the
utilisation of the row.

```{r ead-pools}
m_ead$pools[, .(pool, measure, n, lra, lra_ew, se, moc_est, ccf_dt, ccf_final, ccf_floor, ccf_applied, floor_binding)]
```

`scr_ead_downturn()` takes the periods and a mandatory reason. Under
`"type1"` the downturn value of a pool is `max(lra, observed)`, the
default-weighted realised CCF of the events whose default date falls in
the periods; the applied CCF is recomputed and the ledger records
periods, method and reason.

```{r ead-downturn}
m_ead <- scr_ead_downturn(m_ead, periods = data.frame(start = as.Date("2024-01-01"), end = as.Date("2024-12-01")),
                          reason = "2024 is the stress year of the demo panel")
m_ead$downturn$table
```

## Production, validation and export

`scr_apply()` needs the limit, the drawn amount and the raw drivers of
the admitted set; the utilisation is derived. It returns the pool, the
measure, the applied CCF, the model EAD, the floor EAD
(`drawn + 0.20 * undrawn`), the predicted EAD as the greatest of the
drawn amount, the model and the floor, and whether the floor is the
binding term. The five rows below include an over-limit facility, which
goes to `LF` and where the drawn amount binds, and an undrawn one.

```{r ead-apply}
new_ead <- scr_demo_ead[scr_demo_ead$ref_date == as.Date("2025-06-01") &
                        scr_demo_ead$facility_id %in% c("F0001", "F0002", "F0004", "F0186", "F0615"), ]
new_ead[, c("facility_id", "product", "limit", "drawn")]
ap_ead <- scr_apply(m_ead, new_ead)
ap_ead
```

The floor never binds here because every applied CCF is above 0.20; the
column is there for the pool where it would. The SQL has four blocks:
utilisation and undrawn amount, the driver bin index from the frozen cut
points, the pool from the cells with the `LF` branch, and the applied CCF
with the predicted EAD as a `GREATEST` of the three terms.

```{r ead-sql}
sql_ead <- scr_sql(m_ead, table = "prd.facilities", dialect = "duckdb")
sql_ead_lines <- unlist(strsplit(sql_ead, "\n", fixed = TRUE))
cat(grep("AS pool$", sql_ead_lines, value = TRUE), sep = "\n")
cat(tail(sql_ead_lines, 10), sep = "\n")
```

```{r ead-sql-duckdb, eval = has_duckdb}
con <- suppressMessages(DBI::dbConnect(duckdb::duckdb(), config = list(threads = "1")))
DBI::dbWriteTable(con, "facilities_t", new_ead)
got_ead <- DBI::dbGetQuery(con, paste(scr_sql(m_ead, table = "facilities_t", dialect = "duckdb"), collapse = "\n"))
DBI::dbDisconnect(con, shutdown = TRUE)
got_ead[, c("pool", "ccf_applied", "ead_predicted")]
identical(got_ead$pool, ap_ead$pool)
all.equal(got_ead$ead_predicted, ap_ead$ead_predicted)
```

`scr_ead_validate()` compares realised and predicted on the hold-out per
pool and in total: the one-sided t-test of realised above predicted, the
EAD adequacy ratio (realised over predicted EAD), the gAUC against the
development value, the back-test by cohort and the stability of the pool
and bin distributions. The numeric limits of the lights are a convention
of the package and the summary says so in its last column.

```{r ead-validate}
v_ead <- scr_ead_validate(m_ead)
v_ead
v_ead$summary[, .(test, statistic, p, light)]
v_ead$backtest[, .(cohort, n, realised, predicted, p, adequacy, light_adequacy)]
```

Every light is green because the estimate sits above the realised values
on the hold-out: the margin and the downturn push the applied CCF up, and
an adequacy ratio below one means the predicted EAD covers the realised
one. The workbook carries the funnel, the summaries, the driver bins and
the admission table, the pools and the cells, the downturn and the margin,
the hold-out metrics, the validation blocks, the model card and the
ledger, next to the SQL file.

```{r ead-export, eval = has_openxlsx}
basename(unlist(scr_export(m_ead, out, stamp = FALSE, validation = v_ead)$files))
```

# What both ledgers recorded

Neither parameter can be reconstructed from its pool table alone; the
ledger is the part of the object that says why the numbers are what they
are. The LGD ledger starts in the workout (discounting, cure treatment,
merged defaults, extrapolation, cost allocation, bounds), continues with
the split, the note of each stage and the driver the sign check removed,
the pool merges, the provisional downturn and the final one with its
periods and reason, and the floor with the framework, the asset class and
the secured share.

```{r ledger-lgd}
m$ledger[, .(action, detail, reason)]
```

The EAD ledger has one row per step: how the reference data set was built
(horizon, measure rule, floors, where post-default drawings are booked,
the default level), how the pools were fitted (the admitted drivers, the
margin, the floor arithmetic) and the downturn with its periods and
reason.

```{r ledger-ead}
m_ead$ledger[, .(step, action, reason)]
```

Both ledgers travel unchanged into the `Decision_Ledger` sheet of the
workbook, and the SQL header of each model states the stages, the
downturn status and the floor it was generated with.
