---
title: "PD calibration and rating grades"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PD calibration and rating grades}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The introductory vignette ends with a scorecard whose alignment turns a
score into a probability. That probability is a good ranking device and a
fair estimate of the event rate **of the development sample**; it is not
yet a probability of default (PD) in the sense a capital or provisioning
model needs. Four things separate the two:

* the **default definition**: the target must be a default event built by a
  stated rule (days past due with material arrears, unlikeliness to pay, a
  probation before cure), not whatever flag happened to be in the table;
* the **long-run average**: the level of the PD must be anchored to the
  average one-year default rate observed over many cohorts, not to the
  event rate of one sample, which is a single draw of the cycle;
* **grades**: obligors are pooled into rating grades with a monotone PD,
  enough obligors and defaults to estimate it, and a stable distribution;
* the **margin of conservatism** and the **floor**: the grade PD carries
  an explicit add-on for estimation error and for known deficiencies, and
  it is never reported below the floor of its asset class.

Every step below leaves a record: the merges and repairs of the grades, the
entries of the margin ledger, the parameter table used for the floor. The
names follow the public frameworks (the Basel Framework of the BCBS, the
EBA guidelines on PD and LGD estimation, Resolução BCB 303/2023) only where
they explain a name; the package is a technical tool and the frameworks
are tables of numbers it reads, not prose it interprets.

```{r setup, message = FALSE}
library(scorecraft)
library(data.table)
cfg <- scr_config(verbose = FALSE, nthread = 1, use_ranger = FALSE,
                  use_lightgbm = FALSE, n_boot = 20)
```

## 1. The default flag and the default-rate series

`scr_demo_panel` is a monthly panel: 600 obligors over 36 months with days
past due, arrears, exposure, a restructuring flag and a behavioural score.
`scr_default()` runs the definition of default over it as a small state
machine per obligor: a trigger (`dpd >= 90` with arrears above both the
absolute and the relative materiality thresholds) opens an event, the event
closes after three consecutive months without a trigger (twelve when the
obligor was restructured during the event). The row-level flags are the
product; the events table and the ledger document what the rule did.

```{r default}
d <- scr_default(scr_demo_panel, id = "id", date = "ref_date", dpd = "dpd",
                 arrears = "arrears", exposure = "exposure",
                 restructured = "restructured", config = cfg)
d
head(d$events, 4)
d$ledger$detail
```

`scr_default_rate()` turns the flagged panel into one-year default rates by
cohort: at every cohort start the non-defaulted obligors form the
population and the outcome is a default within the next twelve months. The
long-run average is the arithmetic mean of the cohort rates; the benchmark
is the larger of the last five years' mean and the whole period's mean,
and a flag is raised when the average sits below it. The panel covers less
than two years of complete windows, so the print says plainly that the
average is not a long-run one yet.

```{r default-rate}
dr <- scr_default_rate(d, by = "quarter", config = cfg)
dr
dr$table
```

## 2. Calibrating the scorecard to the central tendency

The scorecard comes from `scr_demo` with the fast configuration. Its
hold-out event rate is 14.5 %, well above the 9.6 % long-run average of the
panel: a sample of the development window, not the cycle.

```{r scorecard}
res <- scr_select(scr_demo, "default", config = cfg, drop = c("id", "churn"),
                  date_col = "ref_date")
sc <- scr_scorecard(res)
sc
```

`scr_calibrate()` re-anchors the probability to the central tendency (CT)
without touching the points: it returns a **new** alignment
`(I*, S*)` whose `predict(..., type = "prob")` is the calibrated PD, while
the scorecard keeps its own alignment for the score. Passing the `scr_dr`
as the target uses its long-run average and records where the CT came from.

```{r calibrate}
cal <- scr_calibrate(sc, target = dr)
cal
```

Four methods are available. `"intercept"` is the prior-correction shift of
King and Zeng, refined by a one-dimensional root so that the mean PD equals
the CT exactly; the slope is untouched, so every rank statistic is
untouched. `"logodds_ab"` and `"qmm"` fit both intercept and slope of
`ln(odds*) = a + b ln(odds)`, the first to a target accuracy ratio, the
second to the accuracy ratio implied by the PDs themselves. `"scaling"` is
the proportional rescaling `PD * CT / ybar`, projected back onto a logit
map. All four hit the CT; only the slope differs, and a monotone map of the
ln(odds) never changes the AUC of the hold-out.

```{r calibrate-compare}
ho <- sc$samples$holdout
auc_score <- sc$metrics[sample == "holdout", auc]
compare <- rbindlist(lapply(c("intercept", "logodds_ab", "qmm", "scaling"), function(m) {
  cm <- scr_calibrate(sc, target = dr, method = m)
  p  <- predict(cm$alignment, ho$link, type = "prob")
  data.table(method = m, mean_pd_before = cm$mean_pd_before, mean_pd_after = cm$mean_pd_after,
             slope_ratio = cm$slope_ratio, auc_before = auc_score,
             auc_after = scr_metrics(p, ho$y, ci = FALSE)$auc)
}))
compare
```

## 3. Rating grades

`scr_grades()` cuts the production score into grades whose PD is monotone.
The default construction is `"geometric"`: a master scale is built between
the 1st and the 99th percentile of the calibrated PD, every grade doubling
(or multiplying by the ratio) the PD of the one before, and its PD
boundaries are converted into score cut points through the calibrated
alignment. Grade 1 is the safest, the highest scores under
`higher_is_safer`. Grades with fewer than `min_obligors` obligors or
`min_defaults` defaults are merged with the neighbour of closer default
rate; the sequence of grade PDs is then repaired by pool-adjacent-violators
when it is not monotone. Every merge is recorded in `repairs`.

```{r grades}
gr <- scr_grades(sc, cal, n_grades = 7, min_obligors = 30, min_defaults = 10)
gr
gr$master_scale
gr$repairs
```

Seven grades were asked for and five survive: the two safest grades of the
scale hold too few defaults on the hold-out and are folded into the third.
Concentration is reported as the Herfindahl index of the grade shares, the
coefficient of variation of the shares and the Herfindahl-based `hi`
index, the three numbers the stability test of section 6 compares against.

```{r concentration}
unlist(gr$concentration)
```

A master scale can also be supplied, when the institution already has one.
`scr_master_scale()` builds it from the first and last midpoints (or from
supplied midpoints or bands); `method = "supplied"` grades by its PD bands
and keeps its labels through the merges.

```{r grades-supplied}
ms <- scr_master_scale(0.005, 0.40, n_grades = 8)
ms
gs <- scr_grades(sc, cal, method = "supplied", master_scale = ms, min_defaults = 10)
gs$table[, .(grade, label, score_lo, score_hi, n, defaults, dr, pd_be)]
gs$repairs
```

### The two-pass path: grade PDs from the cohort series

So far `pd_be` is the default rate of the grade on the hold-out, one
sample again. The grade PD should be the long-run average of the grade's
own cohort default rates, and that series can only be built once the
grades exist. Hence two passes: grade the cohort panel with the cut points
of the first call, build the series by grade with `scr_default_rate()`,
and pass it as `dr` to a second call with identical arguments. The cut
points are the same; only the source of `pd_be` changes, and the ledger
says so.

```{r two-pass}
pnl <- merge(d$flags, as.data.table(scr_demo_panel)[, .(id, date = ref_date, score)],
             by = c("id", "date"))
pnl[, grade := predict(gr, score = score)]
drg <- scr_default_rate(pnl, id = "id", date = "date", default = "default",
                        grade = "grade", by = "quarter", config = cfg)
head(drg$table, 5)
gr2 <- scr_grades(sc, cal, n_grades = 7, min_obligors = 30, min_defaults = 10, dr = drg)
identical(gr2$breaks, gr$breaks)
gr2$table[, .(grade, label, n, dr, pd_mean, pd_be, n_series, t_series)]
gr2$ledger[action == "grade_pd_source", detail]
```

## 4. Margin of conservatism, floor and the PD model

`scr_moc()` appends entries to the margin-of-conservatism ledger of the
grades. Category `C`, the general estimation error, is quantified from the
data: `"ci_timeseries"` takes the one-sided `t` interval of the long-run
average over the cohort series, per grade; `"ci_binomial"` and
`"bootstrap"` work on the obligors when no series exists. Categories `A`
(data and methodological deficiencies) and `B` (changes in standards or
environment) are expert quantities: a value in PD units and a non-empty
reason are both mandatory. The ledger is append-only: `A` and `B` entries
accumulate, a new `C` supersedes the previous one, which stays in the
ledger with `active = FALSE`.

```{r moc}
gm <- scr_moc(gr2, "C", method = "ci_timeseries", dr = drg)
gm <- scr_moc(gm, "A", value = 0.001,
              reason = "unlikeliness-to-pay trigger not available before 2024")
gm
```

`scr_pd()` assembles the PD model: `pd_moc = pd_be + A + B + C`, then the
floor of the asset class under the framework of `params`, and
`pd_final = max(pd_moc, floor)`. `scr_irb_params()` is a table of numbers
selected by a preset; the floors are one of its tables, editable, and an
edited table is flagged in the ledger.

```{r pd}
prm <- scr_irb_params("bcb")
prm$pd_floor
pd <- scr_pd(gm, params = prm, asset_class = "retail_other")
pd
```

The floor does not bind on this portfolio, whose safest grade sits above
5 %. The table carries every intermediate column, so a validator can
follow the number from the sample default rate to the final PD:

```{r pd-table}
pd$table[, .(grade, label, n, dr, pd_be, moc_a, moc_c, pd_moc, floor, pd_final, floor_applied)]
```

## 5. Production: R and SQL agree

`scr_apply()` scores new rows with the scorecard and adds the grade, the
calibrated individual PD and the grade PD; `predict()` does the same from
a vector of scores. The production SQL of the PD model is the SQL of the
scorecard plus one block: a `CASE` over the score cut points for the grade
and another for the final PD.

```{r apply}
new <- head(scr_demo, 5)
scr_apply(pd, new)
predict(pd, score = c(480, 560, 640), type = "pd_final")
```

```{r sql}
sql <- scr_sql(pd, table = "customers", dialect = "duckdb")
cat(tail(sql, 6), sep = "\n")
```

When DuckDB is installed the SQL is run on a few hundred rows and compared
with `scr_apply()`: the score, the grade and the final PD are the same
numbers.

```{r sql-duckdb, eval = requireNamespace("duckdb", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE), message = FALSE}
con <- DBI::dbConnect(duckdb::duckdb(), config = list(threads = "1"))
DBI::dbWriteTable(con, "customers", head(scr_demo, 300))
got <- DBI::dbGetQuery(con, paste(sql, collapse = "\n"))
DBI::dbDisconnect(con, shutdown = TRUE)
exp <- scr_apply(pd, head(scr_demo, 300))
head(got[, c("score", "grade", "pd_final")], 3)
all.equal(got$score, exp$score)
identical(as.integer(got$grade), exp$grade)
all.equal(got$pd_final, exp$pd_final)
```

## 6. Validation on the cohort panel

`scr_pd_validate()` runs the standard battery on a monthly panel with the
default flag and the grade at every month, building the cohorts exactly as
`scr_default_rate()` does. The panel is graded with the cut points of the
PD model through `predict()`, the same path the production SQL follows.

* **Calibration**, per grade pooled over cohorts and per cohort: the
  Jeffreys test, the binomial test with its critical count, the normal `z`,
  the traffic light on the Jeffreys p-value; for the portfolio, the same
  tests on the totals, Hosmer-Lemeshow over the grades, the multi-period
  test over the cohort default rates and the Brier score.
* **Discrimination**: AUC, Gini and KS with a bootstrap interval on the
  score, and the `S` statistic against the development AUC.
* **Stability**: the PSI of the grade distribution per cohort against the
  development sample, the migration matrix pooled over the cohorts, and
  the concentration test on the coefficient of variation.

```{r validate}
pnl[, grade := predict(pd, score = score)]
v <- scr_pd_validate(pd, pnl, id = "id", date = "date", default = "default",
                     grade = "grade", score = "score", by = "quarter")
v
```

The summary is one row per test with its statistic, p-value and light; the
overall light is the worst of them.

```{r validate-summary}
v$summary
```

Two reds, and both are readable. The one-sided tests are green on every
grade because the tested PD is `pd_final`, which carries the margin of
conservatism: the observed default rates sit below it, as they should.
Hosmer-Lemeshow is two-sided and charges that same conservatism as
miscalibration, which is why it is reported next to the one-sided tests
and not instead of them. The grade PSI is large because the panel is a
different population from the development sample: its behavioural score
puts two thirds of the obligors in grade 1 where the hold-out had a
quarter. That is exactly what the PSI is there to flag; a validator would
now ask whether the cut points belong on that population at all.

```{r validate-cohorts}
v$portfolio[, .(cohort, n, d, pd, dr, p_jeffreys, light)]
head(v$stability$psi, 3)
```

### Migration

`scr_migration()` counts the obligors in grade `i` at the cohort start and
grade `j` twelve months later, with defaults and closed accounts in their
own columns, and reports the upper and lower matrix-weighted bandwidths,
the `z` of every off-diagonal cell against its neighbour closer to the
diagonal, and the mobility summary. The validation pools it over the
cohorts whose end date is observed. On this panel the matrix is diagonal:
the behavioural score of `scr_demo_panel` is constant per obligor, so an
obligor either stays in its grade or defaults, and the bandwidths are
undefined for want of a single move.

```{r migration}
v$stability$migration
```

On a panel where grades move, the same function on two grade vectors gives
the full reading:

```{r migration-demo}
set.seed(2)
g0 <- pnl[date == as.Date("2023-01-01"), grade]
g1 <- pmin(5L, pmax(1L, g0 + sample(c(-1L, 0L, 0L, 0L, 1L), length(g0), TRUE)))
g1[sample(length(g1), 25)] <- NA
scr_migration(g0, g1, K = 5)
```

### The point-in-time view

The grade PDs above are through-the-cycle: the long-run average with a
margin. `scr_pd_pit_ttc()` is Vasicek's one-factor bridge between that PD
and the conditional PD of a given state of the systematic factor `z`
(negative for a stressed year, positive for a benign one), with the asset
correlation `rho`; the map is invertible. `scr_pd(philosophy = "pit")`
applies it before the floor.

```{r pit}
rho <- scr_irb_params("bcb")$correlation$retail_other[["lo"]]
pit <- scr_pd_pit_ttc(pd$table$pd_final, z = -1, rho = rho)
data.table(grade = pd$table$grade, pd_ttc = pd$table$pd_final, pd_pit_stressed = pit,
           back_to_ttc = scr_pd_pit_ttc(pit, z = -1, rho = rho, to = "ttc"))
pd_pit <- scr_pd(gm, params = prm, philosophy = "pit", rho = rho, z = -1)
pd_pit$table[, .(grade, pd_ttc, pd_pit, pd_final)]
```

## 7. Deliverables

`scr_export()` writes the PD workbook and the SQL file. With the
validation supplied the workbook carries the test sheets; without it the
sheets exist with an availability row, never a fabricated number.

```{r export, eval = requireNamespace("openxlsx", quietly = TRUE)}
out <- file.path(tempdir(), "scorecraft-pd-vignette")
ex <- scr_export(pd, out, stamp = FALSE, validation = v)
basename(unlist(ex$files))
openxlsx::getSheetNames(ex$files$pd)
```

## 8. What the ledger recorded

Nothing in the PD model depends on a number that was typed into a script
and forgotten. The margin ledger keeps every entry with its category,
method, value and reason, active or superseded; the grade repairs keep
every merge with the count that forced it; the decision ledger of the
model strings the stages together.

```{r ledger}
pd$moc_ledger[, .(id, category, method, level, grade, value, reason, active)]
pd$repairs
pd$ledger[, .(action, detail)]
```
