---
title: "From database to production"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From database to production}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The introductory vignette walks the modelling pipeline on the bundled
`scr_demo` table. This one covers what surrounds it in a real deployment:
pulling the development table out of a database with server-side sampling,
choosing a cut-off and a band policy, moving the scorecard into the database
as SQL and proving that the SQL gives the same numbers as R, watching the
scorecard drift over time, and packaging the deliverables for a validator.

Everything below runs on `scr_demo`, which carries the defects real data has
(sentinel `-999`, missing values, a column that degrades in the last period)
so that each stage has something to show. The configuration is the fast one
used by the examples: one thread, two consensus voters, a short boosting
schedule and a small bootstrap.

```{r setup, message = FALSE}
library(scorecraft)
library(data.table)
cfg <- scr_config(verbose = FALSE, nthread = 1, use_ranger = FALSE,
                  use_lightgbm = FALSE, xgb_rounds = 60, n_boot = 20)
```

## 1. Fetching the development table from a database

In production the table lives in a warehouse, not in a CSV. `scr_connect()`
opens a DBI connection: with a `dsn` it goes through ODBC (and forces
`bigint = "numeric"`, so that a BIGINT column is not mistaken for a
high-cardinality categorical); with a `driver` it accepts any DBI driver,
which is how the database path is exercised here without a DSN.

```{r dsn, eval = FALSE}
# a real deployment: an ODBC data source configured on the machine
con <- scr_connect("my_dsn", timeout = 30)
```

For the vignette we load `scr_demo` into an in-memory SQLite database.
SQLite has no date type, so `ref_date` is written as text; the column is not
a candidate and is only carried along.

```{r sqlite-load, eval = requireNamespace("RSQLite", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)}
con <- scr_connect(driver = RSQLite::SQLite(), dbname = ":memory:")
d <- scr_demo
d$ref_date <- as.character(d$ref_date)
DBI::dbWriteTable(con, "dtm", d)
DBI::dbListTables(con)
```

### Server-side sampling

`scr_fetch()` samples **on the server**. Pulling every row only to discard
most of them in R pays the network cost twice, so the sampling fraction
becomes a `WHERE` clause on a uniform random expression. The expression
follows the connection class: `rand(seed)` on Spark, Databricks and MySQL,
`random()` on PostgreSQL, Redshift and DuckDB, and an integer-modulo
expression on SQLite, whose `random()` returns a signed 64-bit integer and
cannot be seeded. `sample_expr` overrides the choice for a dialect the
package does not know. The query is echoed while messages are on:
`scr_fetch()` follows `scr_verbose()` unless its own `verbose` argument
says otherwise, and `scr_run()` follows `config$verbose`.

```{r fetch, eval = requireNamespace("RSQLite", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)}
half <- scr_fetch(con, "dtm", sample_frac = 0.5, seed = 42)
class(half)
nrow(half)
```

`max_rows` is a memory guard rather than a second sampler. When it binds, the
function counts the table first, reduces the fraction on the server so that
the expected number of rows fits under the cap, and reports the reduction:

```{r fetch-cap, eval = requireNamespace("RSQLite", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)}
capped <- scr_fetch(con, "dtm", max_rows = 1000)
nrow(capped)
```

### Several targets straight from the connection

`scr_run()` chains fetch and selection for a list of targets. The table name
accepts a `{target}` placeholder for one-table-per-target layouts; without
it the same table serves every target. Sibling targets are removed from the
candidate list automatically, and a failure on one target is recorded in the
run set while the loop continues. Two details matter for what follows:

* `scr_run()` records the table name in the configuration, so the SQL
  generated later already points at `dtm` rather than at the placeholder
  `your_table`;
* its `date_col` defaults to `config$oot_date_col`, which is `NULL` here,
  and `ref_date` is dropped from the candidates, so the split is random by
  row. For an out-of-time split pass `date_col = "ref_date"` (and keep the
  column out of `drop`), or fetch the table and call `scr_select()` as in
  section 2.

```{r run, eval = requireNamespace("RSQLite", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)}
rs <- scr_run(con, "dtm", targets = c("default", "churn"), config = cfg,
              drop = c("id", "ref_date", "default", "churn"))
rs
scr_compare(rs)[, .(target, rows, approved, best_model, auc, auc_lo, auc_hi, ks)]
rs$default$config$sql_table
DBI::dbDisconnect(con)
```

## 2. Selection and scorecard with an out-of-time split

The rest of the vignette works on one target, `default`, with the split cut
by whole periods of `ref_date`: the last periods form the hold-out, which is
what the cut-off sweep, the monitoring and the vintage tables need.

```{r select}
res <- scr_select(scr_demo, "default", config = cfg, drop = c("id", "churn"),
                  date_col = "ref_date")
res$split$cutoff
sc <- scr_scorecard(res)
sc
```

## 3. Cut-off strategy

### The sweep

`scr_cutoff()` evaluates candidate cuts. The candidates are quantiles of the
score **on train**, and they are applied **frozen** to the hold-out: both
samples answer the same question at the same score, so a difference between
them measures the stability of the decision, not a difference in the
sampling of the cuts.

```{r cutoff}
ct <- scr_cutoff(sc, n_cuts = 8)
ct
```

How to read one row:

* `cut` is the score; the **safe side** is the high-score side under
  `higher_is_safer` (credit) and the low-score side under
  `higher_is_riskier` (fraud, propensity). The header of the print says
  which.
* `%safe` (`pct_safe`) is the approval rate at that cut; `ev.safe` and
  `ev.risky` are the event rates on each side.
* `ev.avoid` (`events_avoided_pct`) is the share of all events that fall on
  the risky side, the events the cut would decline; `nonevents_lost_pct`,
  in the table, is the share of good accounts declined with them.
* `KS` at the cut is the distance between those two shares. The cut with
  the largest KS separates the populations best, which is not the same as
  the cut that the business should choose.

The table has one row per sample and cut, so the frozen comparison is a
direct read:

```{r cutoff-frozen}
ct$table[cut == ct$cuts[4],
         .(sample, cut, pct_safe, event_rate_safe, events_avoided_pct, ks_at_cut)]
```

### The strategy table

`scr_strategy()` turns the bands (the deciles frozen on train, by default)
into a policy with an explicit economic reading. For each band it computes
the expected profit per account,

$$EP = (1 - p)\,\text{revenue\_good} - p\,\text{loss\_bad},$$

and the break-even event rate, where $EP = 0$, is
`revenue_good / (revenue_good + loss_bad)`. With a revenue of 1,080 per good
account and a loss of 4,500 per bad one, break-even is 19.35 %. The
automatic decision approves a band whose rate is below break-even, sends to
review a band up to 25 % above it and declines the rest; `decisions` fixes
the policy by hand when the business has one.

```{r strategy}
st <- scr_strategy(sc, revenue_good = 1080, loss_bad = 4500)
st
st$breakeven
```

The reading that matters is the band **profitable at the margin**: the
lowest approved band carries an event rate above the portfolio average yet
still returns a positive expected profit per account, because the revenue on
its good accounts covers the loss on its bad ones. Declining it
would look prudent and cost money. The cumulative columns give the
portfolio result of approving everything down to a band:

```{r strategy-cum}
st$table[, .(band, event_rate = round(event_rate, 4), decision,
             ep_per_account = round(ep_per_account, 1), cum_pct = round(cum_pct, 3),
             cum_event_rate = round(cum_event_rate, 4), cum_profit)]
```

### Honest reject inference

A scorecard is fitted on accounts that were accepted and therefore have an
outcome. `scr_reject()` refuses to fabricate one number for the accounts
that have none. Instead it declares the **population scope**, measures the
**coverage per band** and reports a **sensitivity band**: the event rate each
band, and the total, would have if the population without an outcome were
2, 4 or 8 times worse than the observed one. Without a through-the-door
population, the scope is the development sample and the band is flat:

```{r reject-plain}
rj <- scr_reject(sc)
rj
```

With a through-the-door population, the `accepted` flag marks the rows
whose outcome is known. Here `scr_demo` plays the through-the-door
population and only the hold-out rows are treated as accepted, so two thirds
of the population enter through the sensitivity band alone:

```{r reject-population}
acc <- seq_len(nrow(scr_demo)) %in% res$split$holdout_idx
rj2 <- scr_reject(sc, population = scr_demo, accepted = acc)
rj2
rj2$coverage
rj2$sensitivity[band == "TOTAL", .(multiplier, n_dev, events_dev, rate_dev,
                                   n_unknown, events_implied, rate_implied)]
```

How to read it:

* `coverage` is the share of the band's population that has an outcome;
  `coverage_flag` marks bands with no outcome at all (`no_outcome`) and
  bands with fewer than 30 events (`few_events`), where the observed rate
  is itself fragile. The safe bands are flagged here because a 14 % event
  rate leaves few events at the top of the score.
* `rate_unknown` is the assumed rate of the unknown rows in a band
  (`rate_dev` times the multiplier, capped at one), `events_implied` adds
  the implied events to the observed ones, and `rate_implied` is the
  resulting rate over the whole band.
* The `TOTAL` rows are the headline: the observed rate and what it becomes
  under each declared multiplier. The analyst reads the band and states the
  assumption; the package does not choose it.

## 4. Scoring in R

`scr_apply()` materialises in R exactly what the production SQL does: the
frozen pre-processing (training median for missing and sentinel values,
`"MISSING"` for absent categories), the frozen bins and the points. Nothing
is refitted. `what = "all"` returns the logit, the model probability, the
exact score (`a + b * logit`), the whole-points score and, per variable, the
points and the WOE.

```{r apply}
new <- head(scr_demo, 5)
scored <- scr_apply(sc, new, what = "all")
scored[, .(link, prob, score, score_points, vl_score_01_points, vl_score_01_woe)]
```

`scr_reasons()` gives the adverse-action reasons: the variables whose
points fell furthest below their reference, the mean points of the variable
on the training population by default (`reference = "max"` uses the best
bin instead). It only exists for the additive scorecard.

```{r reasons}
scr_reasons(sc, new, k = 3)
```

## 5. Production SQL, and proof that it matches R

### What is generated

`scr_sql(res)` emits the WOE transformation of the approved variables in two
blocks: a CTE `base_scr` that reproduces the pre-processing (imputation by
the **training** median, special-population flags, `COALESCE` of the
categorical missing) and the WOE/BIN `CASE` expressions built from the
authoritative cut points at full precision. Without the first block the WOE
would be applied to data different from what was binned. The vector returned
by `scr_sql()` keeps each block as it was emitted, so the lines are split
here to show a window; `file` writes the complete script.

```{r sql-woe}
sql_woe <- unlist(strsplit(scr_sql(res), "\n", fixed = TRUE))
cat(head(sql_woe, 24), sep = "\n")
i <- grep("AS vl_score_01_bin", sql_woe, fixed = TRUE)
cat(sql_woe[(i + 1):(i + 11)], sep = "\n")
```

`scr_sql(sc)` adds a third block for the scorecard: a CTE `woe_scr` with the
WOE and the bin index, then the final `SELECT` with `score` (exact, from the
WOE), one `<variable>_points` column per variable and `score_points` (the
base points plus the whole points per bin). `table` and `dialect` override
the configuration; the dialects available are `ansi`, `databricks`,
`spark`, `hive`, `mysql`, `mariadb`, `sqlserver`, `bigquery`, `postgres`,
`oracle`, `snowflake`, `redshift`, `duckdb` and `sqlite`. `file` writes the
script to disk.

```{r sql-score}
sql_sc <- unlist(strsplit(scr_sql(sc, table = "prd.customers", dialect = "databricks"),
                          "\n", fixed = TRUE))
cat(head(sql_sc, 12), sep = "\n")
# block 3: the exact score from the WOE columns ...
i <- max(which(sql_sc == "SELECT"))
cat(sql_sc[i:(i + 15)], sep = "\n")
# ... the whole points, and one of the CASE expressions behind them
cat(grep("AS score_points", sql_sc, value = TRUE), sep = "\n")
cat(grep("AS vl_score_01_points", sql_sc, value = TRUE), sep = "\n")
```

### Equivalence, demonstrated

The claim that R and SQL agree is checked by the package tests; it is also
cheap to demonstrate. DuckDB runs in-process, accepts `scr_demo` as is
(dates included) and executes the `duckdb` dialect. The WOE from
`scr_sql(res)` is compared with `scr_apply(res)`, and the score, the whole
points and the WOE columns from `scr_sql(sc)` with `scr_apply(sc)`.

```{r duckdb, eval = requireNamespace("duckdb", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE), message = FALSE}
con <- DBI::dbConnect(duckdb::duckdb(), config = list(threads = "1"))
DBI::dbWriteTable(con, "scr_demo", scr_demo)

# selection SQL: WOE and bin of the approved variables
got_woe <- DBI::dbGetQuery(con, paste(scr_sql(res, table = "scr_demo", dialect = "duckdb"),
                                      collapse = "\n"))
exp_woe <- scr_apply(res, scr_demo, what = "both")
woe_ok <- vapply(scr_selected(res), function(f)
  isTRUE(all.equal(got_woe[[paste0(f, "_woe")]], exp_woe[[paste0(f, "_woe")]])) &&
  identical(got_woe[[paste0(f, "_bin")]], exp_woe[[paste0(f, "_bin")]]), logical(1))
all(woe_ok)

# scorecard SQL: score, whole points and WOE
got_sc <- DBI::dbGetQuery(con, paste(scr_sql(sc, table = "scr_demo", dialect = "duckdb"),
                                     collapse = "\n"))
exp_sc <- scr_apply(sc, scr_demo, what = "all")
all.equal(got_sc$score, exp_sc$score)
all.equal(got_sc$score_points, exp_sc$score_points)
all(vapply(sc$features, function(f)
  isTRUE(all.equal(got_sc[[paste0(f, "_points")]], exp_sc[[paste0(f, "_points")]])), logical(1)))
DBI::dbDisconnect(con, shutdown = TRUE)
```

`score` agrees to floating-point precision, `score_points` and every
`<variable>_points` column agree exactly. The same check on SQLite, whose
arithmetic is less forgiving, is part of the test suite.

## 6. Monitoring

`scr_monitor()` scores any new table with the frozen scorecard and
recomputes, per period of `date_col`, the score PSI against the training
distribution with frozen bands, the CSI of every variable with frozen bins
together with the signed points shift and, when the target is present, the
performance by vintage. It schedules nothing: the analyst calls it when
needed. Here the new data are `scr_demo` itself, whose first four periods
are the training vintages and the last two the hold-out.

```{r monitor}
mo <- scr_monitor(sc, scr_demo, date_col = "ref_date", target = "default", n_boot = 20)
mo
```

### The PSI timeline and its two thresholds

```{r monitor-psi}
mo$psi
```

Every PSI carries **both** thresholds. `flag_fixed` uses the market
convention, 0.10 for a moderate shift and 0.25 for action, which has no
published authority behind it and ignores how many rows produced the
number. `critical` is the sample-size-adjusted critical value at level
`alpha`: under no shift, the PSI over $B$ bands computed from $n$ base and
$m$ comparison rows behaves like $(1/n + 1/m)$ times a chi-squared variable
with $B - 1$ degrees of freedom (Yurdakul and Naranjo, 2020), so the
critical value is $(1/n + 1/m)\,\chi^2_{B-1,\,1-\alpha}$. With 2,800
training rows, 700 rows per period and ten bands it is about 0.03: the fixed
0.10 would let a real shift of that size pass, while on a very small
period it would flag noise. `flag_adjusted` is the verdict against
`critical`; the two flags are reported side by side and disagree exactly
when the sample size makes the fixed threshold misleading.

### The CSI and the signed points shift

The CSI applies the same statistic to each variable's bin distribution and
gets the same pair of thresholds. It is unsigned, so it says that a variable
moved but not where the score went. `points_shift` fills that gap: the change
in bin shares weighted by the points of each bin, which is the amount by which
the variable moved the mean score, with its sign, and it is additive across
variables.

`vl_late` is built to degrade in the last period only, and the monitor
finds it: stable for five periods, then a CSI far above both thresholds
and the largest points shift of the run.

```{r monitor-csi}
mo$csi[variable == "vl_late", .(period, n, csi, flag_fixed, critical, flag_adjusted, points_shift)]
mo$csi[period == max(period)][order(-abs(points_shift))][1:5,
       .(variable, csi, flag_fixed, flag_adjusted, points_shift)]
```

### Performance by vintage

With the target present, each period gets its event rate and its AUC, KS and
Gini with a bootstrap interval. The first four vintages were used for
training, so their higher discrimination is expected; the comparison that
matters is between the hold-out vintages and the scorecard's own hold-out
figures, and then between successive production months.

```{r monitor-vintage}
mo$vintage[, .(period, n, events, event_rate = round(event_rate, 4),
               auc = round(auc, 4), auc_lo = round(auc_lo, 4), auc_hi = round(auc_hi, 4),
               ks = round(ks, 4))]
```

`mo$plan` records the monitoring contract that was applied: both sets of
thresholds, `alpha`, the frozen score bands and the source of each
threshold. It is written to the strategy workbook.

## 7. Deliverables

`scr_export()` writes the deliverables to a directory. `stamp = FALSE` writes
in place; the default creates a timestamped subdirectory so that earlier
runs are never overwritten. For the scorecard, the cut-off, strategy,
reject and monitor objects computed above can be passed in so that the
workbooks carry exactly what was read in this session; otherwise they are
recomputed with defaults.

```{r export, eval = requireNamespace("openxlsx", quietly = TRUE), message = FALSE}
out <- file.path(tempdir(), "scorecraft-prod")
res <- scr_export(res, out, stamp = FALSE)
sc  <- scr_export(sc, out, stamp = FALSE, cutoff = ct, strategy = st, reject = rj2, monitor = mo)
basename(unlist(res$files))
basename(unlist(sc$files))
```

Four workbooks are produced between the two calls, plus the SQL scripts and
a Markdown summary:

* **`selection_default.xlsx`** (from the `scr_result`): `01_Funnel`, every
  input column with the stage it left at and why; `02_Gains`;
  `03_Screening`; `04_Holdout`, the revalidation with frozen bins;
  `05_Models`, `06_Votes` and `07_Consensus`; `08_Ledger`, the
  pre-processing decisions that the SQL reproduces; `09_Redundancy`; and,
  when a coarse-classing lab was applied, `10_Coarse_Classing` and
  `11_Decision_Ledger`.
* **`scorecard_default.xlsx`**: `Score_Summary` (scale, direction and
  `odds_orientation`), `Final_Scorecard` (the points table),
  `Coefficients`, `Sign_Check`, `Alignment` and `Alignment_Bands` (the
  log-odds regression behind the scale), `Model_Card`, and `Challenger`
  and `Swap_Set` when a challenger was fitted.
* **`validation_default.xlsx`**: `Score_Gains_Frozen`, `Variable_Gains_IV`,
  `Discrimination_CI`, `Stability_PSI_Timeline` and
  `Stability_CSI_Timeline` (the monitor tables above), `Stability_Variables`,
  `Calibration` and `Calibration_Bands`, `Performance_By_Vintage` and
  `Rank_Order_Diagnostics`.
* **`strategy_default.xlsx`**: `Population_Scope` and `Band_Coverage` from
  reject inference, `Cutoff_Sweep`, `Strategy_Bands`, `Reject_Sensitivity`
  and `Monitoring_Plan`.

`sql_woe_default.sql` and `sql_score_default.sql` hold the two SQL scripts
of section 5, and `summary_default.md` the executive summary of the
selection.

The writer is hardened for documents that leave the analyst's hands. Every
sheet is sanitised first: factors and list columns become text, and a cell
starting with `=`, `+`, `-` or `@` is prefixed with an apostrophe so that a
spreadsheet cannot interpret it as a formula. The workbook is written to a
temporary file, reopened, and its sheet names and row counts are verified
against what was meant to be written before it is renamed into place; a
verification failure raises an error rather than leaving a partial file. A
sheet with nothing to report receives an `availability` and `reason_code`
row, never a fabricated zero.

## 8. Production checklist

Before the scorecard goes live:

1. **Scope.** The population scope statement from `scr_reject()` is written
   down next to the scorecard, with the coverage per band and the
   multiplier the business is prepared to defend.
2. **Split.** The development split is out-of-time (`date_col` given to
   `scr_select()`), and the hold-out performance in the model card is the
   figure quoted, not the training one.
3. **Cut-off.** The cut and the band policy come from `scr_cutoff()` and
   `scr_strategy()` with the revenue and loss figures the business signed
   off, and the marginal band was read rather than declined by instinct.
4. **SQL.** The script in production is the one written by `scr_export()`
   for the target dialect, with `table` set to the real source; the
   pre-processing CTE was not edited by hand.
5. **Equivalence.** The scored output of the production SQL was compared
   with `scr_apply()` on the same rows, as in section 5, after deployment
   and after any change to the source table.
6. **Reason codes.** `scr_reasons()` is the source of adverse-action
   reasons, with the reference (`mean` or `max`) stated in the policy.
7. **Monitoring contract.** `mo$plan` is filed with the deliverables: both
   PSI thresholds, `alpha`, the frozen bands and the minimum events per
   period below which no verdict is issued.
8. **Cadence.** `scr_monitor()` runs on every new period; a score PSI above
   the adjusted critical value, a CSI shift with a material points shift,
   or a vintage AUC outside the development interval triggers a review
   before a refit.
9. **Deliverables.** The four workbooks, the SQL and the summary are
   archived from a stamped export (`stamp = TRUE`), never overwritten.

## Reference

Yurdakul, B. and Naranjo, J. (2020). Statistical properties of the
population stability index. *Journal of Risk Model Validation*, 14(4),
89-100.
