---
title: "scorecraft: from raw table to aligned scorecard and production SQL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scorecraft: from raw table to aligned scorecard and production SQL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks the whole pipeline on the bundled `scr_demo` table, a
synthetic dataset fabricated with the defects real data has: sentinel
`-999`, missing values, constants, duplicates, a redundant pair, high
cardinality, pure noise and a column that only degrades in the last period.

```{r setup}
library(scorecraft)
cfg <- scr_config("moderate", objective = "risk", verbose = FALSE, nthread = 1,
                  use_ranger = FALSE, xgb_rounds = 60, n_boot = 30)
cfg
```

## Stages 0 to 4: selection

`scr_select()` chains the split (out-of-time by whole periods), the triage,
the optimal binning with screening and hold-out revalidation, and the
multi-strategy consensus. Each stage is also exported on its own
(`scr_split()`, `scr_triage()`, `scr_bin()`, `scr_model()`).

```{r select}
res <- scr_select(scr_demo, "default", config = cfg, drop = "id", date_col = "ref_date")
res
```

The funnel is the deliverable: every input column, the stage it died at and
why.

```{r funnel}
table(scr_funnel(res, cols = "all")$exit_stage)
head(scr_funnel(res, only_selected = TRUE)[, .(feature, total_iv, iv_holdout, ks, psi, psi_flag_adjusted)])
```

## Stages 4 and 5: scorecard and alignment

`scr_scorecard()` fits the logistic regression on the WOE columns, checks
the signs, and aligns the logit to the declared scale by regressing
empirical log-odds on score bands and composing with the PDO map. The
alignment object records the odds orientation so that two scorecards
aligned this way are comparable.

```{r scorecard}
sc <- scr_scorecard(res)
sc
sc$alignment
head(sc$points[, .(variable, bin, woe, coef, points)])
```

Discrimination always comes with a bootstrap confidence interval, and the
score PSI with both thresholds.

```{r metrics}
scr_score_metrics(sc)[, .(sample, auc, auc_lo, auc_hi, ks, gini)]
sc$stability$score
```

## Stage 6: cut-off, strategy and reject inference

```{r strategy}
scr_cutoff(sc, n_cuts = 8)
scr_strategy(sc, revenue_good = 1080, loss_bad = 4500)
scr_reject(sc)
```

## Production: R and SQL agree

```{r apply}
new <- head(scr_demo, 5)
scr_apply(sc, new)
scr_reasons(sc, new, k = 2)
```

```{r sql}
cat(tail(scr_sql(sc, table = "prd.customers", dialect = "databricks"), 8), sep = "\n")
```

## Monitoring

`scr_monitor()` recomputes the score PSI, the per-variable CSI with the
signed points shift and the performance by vintage on any new data. It never
schedules itself.

```{r monitor}
scr_monitor(sc, scr_demo, date_col = "ref_date", target = "default", n_boot = 30)
```

## Deliverables

`scr_export()` writes hardened `.xlsx` workbooks (selection, scorecard,
validation, strategy), the SQL files and a Markdown summary.

```{r export, eval = requireNamespace("openxlsx", quietly = TRUE)}
out <- file.path(tempdir(), "scorecraft-vignette")
basename(unlist(scr_export(sc, out, stamp = FALSE)$files))
```
