---
title: "Expected loss and regulatory capital"
description: >
  Expected loss, the IRB risk-weight function, input and output floors,
  the foundation approach, EL against provisions, capital SQL and
  accounting expected credit loss.
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

This is the third of three articles on the IRB risk parameters. [PD
calibration and rating grades](pd-calibration-and-grades.html) builds the
probability of default and [LGD and EAD under IRB](lgd-and-ead-under-irb.html)
the loss given default and the exposure at default, both starting from
the scorecard pipeline of [Get
started](https://evandeilton.github.io/scorecraft/articles/scorecraft.html).
This article turns the three parameters into expected loss, IRB risk
weights, capital with the output floor and accounting ECL. It works on the
bundled `scr_demo_portfolio`, whose PD, LGD and EAD columns are of the
kind those models produce (grade PDs, pool LGDs, exposures), rather than
on the objects built there, so that the portfolio can cover six asset
classes at once.

```{r setup, message = FALSE}
library(scorecraft)
library(data.table)
cfg <- scr_config(verbose = FALSE, nthread = 2)
```

## 1. Expected and unexpected loss

The loss a portfolio produces in an average year is its **expected loss**,
`EL = PD * LGD * EAD`; it is a cost of doing business and is covered by
provisions and pricing. The loss in a bad year exceeds that average, and
the gap between a high quantile of the loss distribution and its mean is
the **unexpected loss** that capital must absorb. The IRB risk-weight
function turns a through-the-cycle PD into the PD of a year at the 99.9% quantile
of a one-factor model, and charges capital for the difference
between the loss at that quantile and the expected loss. Every number that
a regime fixes in that calculation (PD and LGD floors, asset correlations,
the maturity rule, the confidence level, the standardised risk weights
that the output floor compares against) lives in a table returned by
`scr_irb_params()` and selected by a preset; nothing is hard-coded in the
functions that consume it, and an edited table is recorded as such on
every object built from it.

## 2. The parameter tables

`scr_irb_params()` returns every regime-specific number as a table; the
print summarises them.

```{r params}
p <- scr_irb_params("bcb")
p
p$pd_floor
```

There are three presets. `"bcb"` takes its values from BCB Resolution
303/2023, `"basel3_final"` from the consolidated Basel Framework, whose
CRE31 chapter is the risk-weight function, and `"crr3"` from the EU
Capital Requirements Regulation as amended in 2024; `"crr3"` currently
carries the same tables as `"basel3_final"`. Between `"bcb"` and
`"basel3_final"` the visible differences are the supervisory LGD table of
the foundation approach (the `"bcb"` preset carries 75% for a senior
unsecured claim and 45% for a claim with priority; the Basel table
separates corporates, 40%, from financial institutions, 45%) and the
firm-size bounds of the SME correlation adjustment. Each table is a
default to be checked against the text in force, not a citation.

```{r params-diff}
p_b3 <- scr_irb_params("basel3_final")
rbind(cbind(framework = "bcb", p$lgd_firb), cbind(framework = "basel3_final", p_b3$lgd_firb))
rbindlist(list(c(framework = "bcb", p$correlation$sme), c(framework = "basel3_final", p_b3$correlation$sme)))
```

Any cell can be edited and the object passed back through `params`; the
function that receives it flags `params_modified` in its model card.

## 3. One exposure: `scr_el()` and `scr_irb_rw()`

Expected loss is the primitive. Defaulted exposures replace `PD * LGD`
with the best estimate of expected loss, `ELBE`.

```{r el}
scr_el(c(0.01, 0.02), 0.45, c(1000, 2000))
scr_el(0.02, 0.45, 1000, defaulted = TRUE, elbe = 0.6)
```

`scr_irb_rw()` returns the intermediate quantities of the risk-weight
function, not only the answer: the PD and LGD after floors, the maturity
after clipping, the correlation `r`, the maturity adjustment `ma` and the
capital requirement `k`. A corporate exposure at PD 1%, LGD 45% and
maturity 2.5 years carries a risk weight of 92.32%.

```{r rw-examples}
scr_irb_rw(0.01, 0.45, m = 2.5, asset_class = "corporate", params = p)
rbind(
  mortgage     = scr_irb_rw(0.01, 0.20, asset_class = "retail_mortgage", params = p),
  qrre         = scr_irb_rw(0.02, 0.80, asset_class = "qrre_revolver", params = p),
  retail_other = scr_irb_rw(0.02, 0.50, asset_class = "retail_other", params = p)
)[, .(pd_used, lgd_used, r, k, rw)]
```

Retail classes have no maturity adjustment (`b` is `NA`, `ma` is one), so
for them `K` is exactly `LGD` times the distance between the stressed and
the unconditional PD. `scr_pd_stress()` is that stressed PD, and at
`q = 0.999` with the regulatory correlation it reproduces `k` to machine
precision.

```{r stress}
o <- scr_irb_rw(0.02, 0.50, asset_class = "retail_other", params = p)
c(k = o$k, by_hand = 0.5 * (scr_pd_stress(0.02, o$r, 0.999) - 0.02))
scr_pd_stress(0.02, rho = 0.15, q = c(0.5, 0.95, 0.99, 0.999))
```

The shape of the function differs by asset class because the correlation
does: fixed for mortgages (0.15) and revolving retail (0.04), decreasing
in PD for corporates and other retail. Plotted at a common LGD of 45%:

```{r rw-curve}
grid <- exp(seq(log(1e-4), log(0.3), length.out = 80))
classes <- c("corporate", "retail_mortgage", "qrre_revolver", "retail_other")
curves <- lapply(classes, function(a) scr_irb_rw(grid, 0.45, asset_class = a, params = p)$rw)
plot(grid, curves[[1]], type = "n", log = "x", ylim = c(0, max(unlist(curves))),
     xlab = "PD (log scale)", ylab = "risk weight", main = "IRB risk weight by PD, LGD 45%")
for (i in seq_along(classes)) lines(grid, curves[[i]], lwd = 2, col = i)
legend("topleft", classes, col = seq_along(classes), lwd = 2, bty = "n")
```

The flat start of every curve is the PD floor: below it the risk weight no
longer falls with the PD. The `floors_hit` attribute counts the rows where
each floor was binding, and `apply_floors = FALSE` shows what the function
would return without them.

```{r floors}
r_floored <- scr_irb_rw(grid, 0.45, asset_class = "retail_other", params = p)
attr(r_floored, "floors_hit")
r_raw <- scr_irb_rw(grid, 0.45, asset_class = "retail_other", params = p, apply_floors = FALSE)
data.table(pd = grid, rw_floored = r_floored$rw, rw_no_floor = r_raw$rw)[pd < 6e-4][c(1, 5, 10, 15)]
# LGD floors: an unsecured own estimate of 10% is lifted to the floor of its class,
# except for the mortgage, whose 5% floor does not bind
scr_irb_rw(0.01, 0.10, asset_class = c("retail_other", "qrre_revolver", "corporate", "retail_mortgage"), params = p)$lgd_used
# secured by real estate, the retail floor is 10% instead of the unsecured 30%
scr_irb_rw(0.01, 0.05, asset_class = "retail_other", collateral = "real_estate", params = p)$lgd_used
```

## 4. The standardised comparison

The output floor compares the IRB result with a fraction of what the
standardised approach would require, so the same book needs a standardised
risk weight per exposure. `scr_sa_rw()` looks it up in `params$sa_rw`:
regulatory retail, mortgages by loan-to-value band, corporates by rating
bucket or the SME weight, defaulted exposures by provision ratio.

```{r sa}
scr_sa_rw(c("retail_other", "retail_mortgage", "retail_mortgage", "corporate", "corporate", "corporate_sme"),
          ltv = c(NA, 0.55, 0.95, NA, NA, NA), rating = c(NA, NA, NA, "A+", NA, NA))
scr_sa_rw("retail_other", defaulted = TRUE, provision_ratio = c(0.1, 0.3))
p$sa_rw[asset_class == "retail_mortgage" & sub_class == "standard"]
```

## 5. The portfolio: `scr_capital()`

`scr_demo_portfolio` holds 5,000 exposures in six segments that map one to
one onto the asset classes; PD is a grade PD and LGD a pool value, so
segment by grade is homogeneous in PD and LGD, while maturity varies
inside the corporate pools (section 6 comes back to this). About 3% of
the rows are in default with an `elbe` and a provision. The whole book is
run under the advanced approach (`capital_approach = "airb"`, the
default) to show every input at work; under the final Basel framework and
CRR3 a book of large corporates would be restricted to the foundation
approach, shown at the end of this section.

```{r capital}
d <- scr_demo_portfolio
table(d$segment, d$asset_class)
cap <- scr_capital(d, segment = "segment", asset_class = "asset_class", m = "m",
                   defaulted = "defaulted", elbe = "elbe", provisions = "provision",
                   ltv = "ltv", rating = "rating", sales = "sales", transactor = "transactor",
                   grade = "grade", id = "id", config = cfg, keep_rows = TRUE)
cap
```

### Totals and the reconciliation by segment

Every figure on the print is in `cap$totals`, and `cap$segments` breaks it
down: EAD-weighted inputs, the IRB and standardised risk-weighted assets
side by side, expected loss and provisions.

```{r totals}
with(cap$totals, data.table(ead, el, el_rate, rwa_irb, rwa_sa, irb_sa_ratio, density, capital))
cap$segments[, .(segment, n, ead, pd_mean, lgd_mean, rw, rwa_irb, irb_sa_ratio, el, provisions, shortfall_excess)]
```

The `irb_sa_ratio` column says where the IRB approach saves the most
against the standardised one; mortgages sit far below the 72.5% line on
their own, the large corporate book just above it and the SME book well
above, and the output floor is
applied to the total, not by segment. The total here is credit risk only:
the regulatory floor compares the whole of the risk-weighted assets, market
and operational risk included, with 72.5% of their standardised
counterparts, so the figure below is a credit-portfolio view of the floor,
not the bank's.

### Floors and the output-floor bridge

`cap$floors` measures each input floor by the rows it binds, their EAD
and the risk-weighted assets it adds. Here only the PD floor bites (the
two safest grades start below it on purpose), and the bridge from the
unfloored IRB figure to the reported one reads:

```{r floors-bridge}
cap$floors
with(cap$totals, data.table(
  step  = c("rwa_irb_no_floors", "input_floors", "rwa_irb", "rwa_sa", "output_floor_rwa", "rwa_reported"),
  value = c(rwa_irb_no_floors, rwa_irb - rwa_irb_no_floors, rwa_irb, rwa_sa, rwa_floor, rwa_reported)))
with(cap$totals, data.table(floor_binding, headroom))
```

`rwa_reported` is the larger of the IRB figure and `output_floor * rwa_sa`;
`headroom` is the distance to the floor, negative when it binds.

### Expected loss against provisions

Regulatory EL is compared with the provision stock. A shortfall is deducted
from common equity tier 1; an excess counts as tier 2 up to 0.6% of the
IRB risk-weighted assets, which is why `tier2_addback` can be smaller than
`excess`. The comparison is made on the totals, defaulted and performing
exposures together. The Basel text (CRE35) and the CRR compare the two
separately, so that an excess of specific provisions on defaulted
exposures cannot cover a shortfall on performing ones; where that
distinction matters, run the function on each part and combine the
results.

```{r provisions}
with(cap$totals, data.table(el, provisions, shortfall, excess, tier2_cap, tier2_addback))
```

### Sensitivity

The grid re-runs the whole function under fixed shocks. The two Vasicek
rows stress every performing PD with `scr_pd_stress()` at the exposure's
own correlation and feed the stressed PD back into the function, so that
the correlation follows the PD; they are the closest the grid comes to a
bad-year capital figure.

```{r sensitivity}
cap$sensitivity
```

### The foundation approach

Under `capital_approach = "firb"` the LGD is the supervisory value of the
claim type, read from `params$lgd_firb` through the `claim` column, no LGD
floor applies and the maturity is fixed at `params$m_default`. The claim
names belong to the preset. Here the large corporates are run as senior
unsecured claims under the two presets and set beside the advanced run of
the same rows.

```{r firb}
big <- d[d$segment == "corporate_large", ]
cap_firb <- function(framework, claim) {
  big$claim <- claim
  scr_capital(big, segment = "segment", asset_class = "asset_class", m = "m",
              defaulted = "defaulted", elbe = "elbe", sales = "sales", rating = "rating",
              grade = "grade", claim = "claim", params = scr_irb_params(framework),
              config = scr_config(verbose = FALSE, nthread = 2, framework = framework,
                                  capital_approach = "firb"))$totals
}
airb  <- cap$segments[segment == "corporate_large"]
f_bcb <- cap_firb("bcb", "senior_unsecured")
f_b3  <- cap_firb("basel3_final", "senior_unsecured_corporate")
data.table(run = c("airb, own LGD", "firb, bcb preset", "firb, basel3_final preset"),
           rwa_irb = c(airb$rwa_irb, f_bcb$rwa_irb, f_b3$rwa_irb),
           el = c(airb$el, f_bcb$el, f_b3$el))
```

The preset drives the result. With the 40% of the Basel table, equal to
the own LGD of this book, the foundation run is lower than the advanced
one because the maturity is fixed at 2.5 years where the book averages
three, and because the defaulted rows carry no `LGD - ELBE` charge under
the foundation approach. With the 75% the `"bcb"` preset carries for a senior
unsecured claim, the risk-weighted assets rise by about 70% and the
expected loss by about a quarter (less than in proportion, because the
defaulted rows keep their `ELBE`).

## 6. The same numbers in SQL

The production query does not need a normal quantile: the constants of
every pool (PD, LGD, correlation, `K`, `RW`, all after floors) are computed
in R and emitted as a `pool_params` CTE, and the exposure table is joined
to it on segment and grade. Defaulted rows use `ELBE * EAD` and
`K = max(0, LGD - ELBE)` from their own columns.

```{r sql}
sql <- scr_sql(cap, table = "portfolio", dialect = "duckdb")
cat(grep("^-- (CTE|  |NOTE)", sql, value = TRUE), sep = "\n")
cat(tail(sql, 21), sep = "\n")
```

The retail pools are homogeneous, so the query reproduces R row by row;
the corporate pools carry a maturity that varies inside the pool, so their
constants are EAD-weighted and the match is exact at pool level (the
header of the SQL says so). Both facts can be checked on DuckDB.

```{r duckdb, message = FALSE}
con <- DBI::dbConnect(duckdb::duckdb())
DBI::dbWriteTable(con, "portfolio", d)
got <- DBI::dbGetQuery(con, paste(sql, collapse = "\n"))
got <- got[match(d$id, got$id), ]
ex <- cap$exposures
retail <- !d$segment %in% c("corporate_large", "corporate_sme")
data.table(id = d$id, segment = d$segment, el_r = ex$el, el_sql = got$el, rwa_r = ex$rwa, rwa_sql = got$rwa)[retail][1:4]
c(el = all.equal(got$el[retail], ex$el[retail]), rwa = all.equal(got$rwa[retail], ex$rwa[retail]))
agg <- DBI::dbGetQuery(con, paste(scr_sql(cap, table = "portfolio", dialect = "duckdb", level = "portfolio"), collapse = "\n"))
agg <- agg[match(cap$segments$segment, agg$segment), ]
c(rwa = all.equal(agg$rwa, cap$segments$rwa_irb), el = all.equal(agg$el, cap$segments$el))
DBI::dbDisconnect(con, shutdown = TRUE)
```

## 7. Accounting expected credit loss: `scr_ecl()`

`scr_ecl()` computes the discrete-time expected credit loss: a
survival-weighted sum of marginal monthly hazards times LGD and EAD,
discounted at the effective interest rate. With a flat hazard, no
discounting and no prepayment the sum collapses to the closed form
`LGD * EAD * (1 - (1 - h)^T)`.

```{r ecl-flat}
cfg_none <- scr_config(verbose = FALSE, nthread = 2, ecl_discount = "none")
h <- 0.01; lgd <- 0.4; ead <- 1000
e1 <- scr_ecl(h, lgd, ead, t_max = 36L, config = cfg_none)
c(ecl_12m = e1$totals$ecl_12m, closed_form = lgd * ead * (1 - (1 - h)^12),
  ecl_life = e1$totals$ecl_life, closed_form = lgd * ead * (1 - (1 - h)^36))
```

On the portfolio, the annual grade PD becomes a flat monthly hazard, run
over a term that depends on the product: twenty years for mortgages, the
contractual maturity for corporates and three years for the other retail
lines. The hazard matrix is zero after the term of each row, so the
lifetime figure stops there. The stage is allocated by the rule: 90 days
past due for stage 3; 30 days past due, or a 12-month PD at least twice
the one at origination, for stage 2. The 30-day backstop is the
rebuttable presumption of IFRS 9; the ratio of two is a common convention,
set by `ecl_sicr_ratio`, not a threshold the standard prescribes. Three
scenarios are weighted. The `z` shock moves every monthly hazard through
the one-factor model with the correlation `rho`, given explicitly here
because the default (0.15) is only a placeholder; applied month by month,
the shock raises the 12-month PD more than the same `z` applied to the
annual PD would. `lgd_add` and `ead_mult` do what their names say.

Two inputs are shortcuts that a real ECL model would not take. The PD
here is a regulatory grade PD, through-the-cycle and possibly carrying a
margin and a floor, and the LGD is a pool value that may include a
downturn component; IFRS 9 asks for unbiased, point-in-time,
forward-looking estimates, such as `pd_be` mapped to the current state of
the cycle (see the point-in-time section of the PD article) and an LGD
without the downturn and the margin. The mechanics below are unchanged by
that choice; the numbers are not.

```{r ecl-portfolio}
hz <- 1 - (1 - d$pd)^(1 / 12)
term <- ifelse(d$segment == "mortgages", 240L,
               ifelse(is.na(d$m), 36L, as.integer(round(12 * d$m))))
hz_term <- outer(hz, rep(1, max(term)))
hz_term[col(hz_term) > term] <- 0
ecl <- scr_ecl(hz_term, d$lgd, d$ead, eir = d$eir, dpd = d$dpd, pd_orig = d$pd_orig,
               rho = 0.10, segment = d$segment, id = d$id,
               scenarios = list(base = list(), downturn = list(z = -1, lgd_add = 0.05), upturn = list(z = 1)),
               weights = c(0.5, 0.3, 0.2), config = cfg)
ecl
ecl$scenarios
ecl$stages
```

Accounting ECL and regulatory EL measure different things and are not
expected to agree: the ECL of stage 2 is a lifetime figure, the ECL of
stage 3 is `LGD * EAD` rather than `ELBE * EAD`, and the scenario weights
tilt the accounting number. Laying the two side by side per segment is
nevertheless the first table a reconciliation asks for.

```{r ecl-vs-el}
merge(cap$segments[, .(segment, ead, el_regulatory = el, provisions)],
      ecl$segments[, .(segment, ecl_accounting = ecl, share_stage2, share_stage3)], by = "segment")[order(-ead)]
c(el_regulatory = cap$totals$el, ecl_accounting = ecl$totals$ecl, provisions = cap$totals$provisions)
```

## 8. Deliverables

`scr_export()` writes one workbook with the summary, the configuration and
every table of the object, plus the SQL file.

```{r export, message = FALSE}
out <- file.path(tempdir(), "scorecraft-capital")
ex <- scr_export(cap, out, stamp = FALSE)
basename(unlist(ex$files))
openxlsx::getSheetNames(ex$files$xlsx)
```

## 9. The ledger and the model card

Every choice the function made on the caller's behalf is a row of the
ledger: the preset and approach, where the inputs came from, how many rows
each floor touched, whether the output floor binds, and the provision
comparison. The model card carries the numbers a validator needs to
reproduce the run, including whether the parameter tables were edited.

```{r ledger}
cap$ledger[, .(action, detail)]
mc <- cap$model_card
keys <- c("framework", "approach", "params_modified", "pd_source", "n_exposures", "n_defaulted",
          "n_pools", "rwa_irb", "rwa_sa", "output_floor_binding", "density", "capital",
          "shortfall", "excess", "scaling_factor", "confidence")
data.table(field = keys, value = vapply(mc[keys], function(v) format(v, digits = 6), character(1)))
```

Two runs with the same framework, approach, `params_modified = FALSE` and
the same input sources are comparable; a run with `params_modified = TRUE`
is a different regime and must say which cells changed.
